---
title: "sra-submission"
output: html_document
date: "2025-01-07"
---

# Anonymise patient ID

```{r}
#reassign randomised patient ID
patients_reassigned <- data.frame("patient_id" = sample(69:1),
                                  "Record.ID" = unique(all_selected_samples_collected_all_datasets$Record.ID))
# patients_reassigned %>% filter(patient_id == Record.ID)
write.csv(patients_reassigned, paste0(preprocessing_path, "sra-submission/patients-reassigned.csv"), row.names = FALSE)
```

# RNAseq

234 total

Use Human 1.0 SRA package

```{r}
# Sample metadata
cols <- data.frame("organism" = "Homo sapiens", 
          "isolate" = NA, 
          "age" = NA,
          "biomaterial_provider" = "Lung Transplant Service, Department of Respiratory Medicine, Alfred Hospital, Melbourne, Victoria, Australia",
          "collection_date" = NA,
          "geo_loc_name" = "Australia",
          "sex" = NA,
          "tissue" = "Lungs",
          "sample_type" = "broncho-alveolar lavage"
          )

r <- cols %>% 
  cross_join(rna_corrected$samples) %>% 
  left_join(patients_reassigned) %>%
  dplyr::select("sample_name" = SampleID,
                "sample_id" = SampleID, 
                colnames(cols),
                "sequencing_batch" = Batch, 
                "patient_id" = patient_id, #scrambled id
                "disease" = Group,
                "disease_stage" = Status,
                "time_post_transplant" = Time_interval_detailed) 

write.csv(r, paste0(preprocessing_path, "sra-submission/rnaseq-samples-metadata.csv"), row.names = FALSE)

identical(rna_corrected$samples$Patient_days, all_selected_samples_collected_all_datasets %>% filter(Dataset == "Transcriptomics") %>% pull(Patient_days))
```

```{r}
# SRA metadata
l <- read.csv(paste0(preprocessing_path, "Transcriptomics/Data/nf-core/samples-sequenced.csv")) %>% 
  filter(SampleID %in% r$sample_name) %>%
  mutate(fastq_1 = gsub("/home/giac0002/of33_scratch/Giulia/nf-core/", "", fastq_1),
         fastq_2 = gsub("/home/giac0002/of33_scratch/Giulia/nf-core/", "", fastq_2),
         library_ID = gsub("_1.fq.gz", "", fastq_1)) %>%
  rename(sample_name = SampleID) %>%
  dplyr::select(!strandedness)
  
r_sra <- data.frame("sample_name" = r$sample_name,
                   "title" = "RNA-seq of Homo sapiens: bronchoalveolar lavage of transplanted lung",
                   "library_strategy" = "RNA-seq",
                   "library_source" = "TRANSCRIPTOMIC",
                   "library_selection" = "RANDOM",
                   "library_layout" = "paired",
                   "platform" = "ILLUMINA",
                   "instrument_model" = "Illumina NovaSeq 6000",
                   "design_description" = NA,
                   "filetype" = "fastq") %>%
  left_join(l) %>%
  rename(filename = fastq_1, filename2 = fastq_2)

write.table(r_sra, paste0(preprocessing_path, "sra-submission/rnaseq-sra-metadata.tsv"), row.names = FALSE, sep = "\t")
```

## Match biosample accession ids

```{r}
# RNA
rna_sample_ids <- rna_corrected$samples %>% 
  dplyr::select("rna_sample_id" = SampleID, Patient_days) %>%
  remove_rownames() %>% 
  mutate(rna_sample_id = ifelse(rna_sample_id == "5134a", "5134", rna_sample_id), 
         Patient_days_lobe = ifelse(Patient_days == "19_300", "19_300_Left", Patient_days)) #add lobe for tx sample, only one, 106, 114 and 116 not in dataset

rna_attributes <- read.delim(paste0(preprocessing_path, "sra-submission/rna-attributes.tsv")) %>% 
  dplyr::select("biosample_accession" = accession, sample_id) %>% 
  rename(rna_sample_id = sample_id) %>%
  mutate(rna_sample_id = as.character(rna_sample_id)) %>%
  left_join(rna_sample_ids)

bact_sample_ids <- data.frame(sample_data(bact), row.names = NULL) %>% dplyr::select(Patient_days_lobe) %>%
  left_join(data.frame(sample_data(bact.filt), row.names = NULL) %>% 
              mutate(Patient_days_lobe = paste(Record.ID, Days, Lobe, sep = "_"),
                     Patient_days_lobe = gsub("_NA", "", Patient_days_lobe)) %>%
              dplyr::select("16s_sample_id" = SampleID, Patient_days_lobe) ) %>%
  mutate(Patient_days = gsub("_Right|_Left", "", Patient_days_lobe))

fungi_sample_ids <- data.frame(sample_data(fungi), row.names = NULL) %>% dplyr::select(Patient_days_lobe) %>%
  left_join(data.frame(sample_data(fungi.filt), row.names = NULL) %>% 
              mutate(Patient_days_lobe = paste(Record.ID, Days, Lobe, sep = "_"),
                     Patient_days_lobe = gsub("_NA", "", Patient_days_lobe)) %>%
              dplyr::select("its_sample_id" = SampleID, Patient_days_lobe) ) %>%
  mutate(Patient_days = gsub("_Right|_Left", "", Patient_days_lobe))

merged_attributes <- rna_attributes %>% full_join(bact_sample_ids) %>% full_join(fungi_sample_ids)

table(is.na(merged_attributes$`16s_sample_id`)) #365
table(is.na(merged_attributes$`its_sample_id`)) #200
table(is.na(merged_attributes$`rna_sample_id`)) #234

#merged_attributes %>% filter(!Patient_days %in% datasets[[1]]$Patient_days)
```

# Microbiome
## 16S

Use MIMARKS: specimen, human-associated

Had to change Human 16S metagenome to Homo sapiens 16S metagenome

365 total + 57 ctrls (all ctrl samples used during preprocessing) = 422

```{r}
# Sample metadata
cols <- data.frame("organism" = "Homo sapiens 16S metagenome", 
          "isolate" = NA, 
          "collection_date" = NA,
          "env_broad_scale" = "Lower human respiratory tract",
          "env_local_scale"	= "not applicable",
          "env_medium" = "not applicable",
          "geo_loc_name" = "Australia",
          "host" = "Homo sapiens",
          "isol_growth_condt" = "not applicable",
          "lat_lon" = "not collected",
          "tissue" = "Lungs"
          )

b <- cols %>% 
  cross_join(data.frame(sample_data(bact), row.names = NULL) %>% 
              dplyr::select(Patient_days_lobe, Record.ID, Time_interval_detailed, Group, Status) ) %>%
  left_join(data.frame(sample_data(bact.filt), row.names = NULL) %>% 
               dplyr::select(SampleID, MiSeqID, Batch, "Patient_days_lobe" = Patient_days) ) %>% 
  left_join(patients_reassigned) %>%
  left_join(merged_attributes %>% dplyr::select(biosample_accession, "SampleID" = `16s_sample_id`)) %>%
  dplyr::select(biosample_accession,
                "sample_name" = SampleID, 
                colnames(cols),
                "dna_extraction_id" = Batch,
                "sequencing_id" = MiSeqID, 
                "patient_id" = patient_id, 
                "disease" = Group,
                "disease_stage" = Status,
                "time_post_transplant" = Time_interval_detailed) %>%
  mutate("age" = NA, sex = NA, "sample_type" = "broncho-alveolar lavage"
)

# include negative and positive controls
ctrls <- cols %>%
  cross_join(data.frame(sample_data(bact.raw))) %>%
  filter(Type != "Sample") %>%
  dplyr::select("dna_extraction_id" = Batch, "sequencing_id" = MiSeqID, "sample_name" = SampleID, "sample_type" = Type, colnames(cols)) %>%
  filter(dna_extraction_id %in% c(unique(b$dna_extraction_id), "PCR"), sequencing_id %in% unique(b$sequencing_id)) %>%
  remove_rownames()

# check overlap
length(unique(ctrls$sequencing_id)) #"33" "40" "42" "48" "25" "23" "27" "32"
length(unique(ctrls$dna_extraction_id)) #"B23" "B24" "B28" "B30" "B17" "B10" "B11" "B12" "B13" "B5"  "B15" "B16" "B18" "B19" "B2"  "B20" "B21" "B22" "B25" "B26" "B27" "B29" "B3" "B4"  "B6"  "B7"  "B8"  "B9" "PCR

length(intersect(unique(ctrls$sequencing_id), unique(b$sequencing_id))) #8 matches
length(intersect(unique(ctrls$dna_extraction_id), c(unique(b$dna_extraction_id), "PCR"))) #29 matches

b <- b %>% full_join(ctrls)

write.table(b, paste0(preprocessing_path, "sra-submission/16s-samples-metadata.tsv"), row.names = FALSE, sep = "\t")

identical(sample_data(bact)$Patient_days, all_selected_samples_collected_all_datasets %>% filter(Dataset == "Bacteria") %>% pull(Patient_days))
```

```{r}
# SRA metadata
b_sra <- data.frame("sample_name" = b$sample_name,
                   "library_ID" = b$sample_name,
                   "title" = "16SrRNA sequencing of Homo sapiens: bronchoalveolar lavage of transplanted lung",
                   "library_strategy" = "AMPLICON",
                   "library_source" = "METAGENOMIC",
                   "library_selection" = "RANDOM",
                   "library_layout" = "paired",
                   "platform" = "ILLUMINA",
                   "instrument_model" = "Illumina Miseq",
                   "design_description" = NA,
                   "filetype" = "fastq") %>%
  left_join(b) %>%
  mutate(filename = paste(sample_name, "R1.fastq.gz", sep = "-"), 
         filename2 = paste(sample_name, "R2.fastq.gz", sep = "-"))  

table(duplicated(b_sra_sub1$biosample_accession))
  
b_sra_sub1 <- b_sra %>% filter(!is.na(biosample_accession)) %>% dplyr::select(!sample_name) #cannot have both biosample accession and sample name
b_sra_sub2 <- b_sra %>% filter(is.na(biosample_accession)) %>% dplyr::select(!biosample_accession)

write.table(b_sra_sub1, paste0(preprocessing_path, "sra-submission/16S-sra-metadata-sub1.tsv"), row.names = FALSE, sep = "\t") #201 > 202
write.table(b_sra_sub2, paste0(preprocessing_path, "sra-submission/16S-sra-metadata-sub2.tsv"), row.names = FALSE, sep = "\t") #221 > 220
write.table(b_sra, paste0(preprocessing_path, "sra-submission/16S-sra-metadata.tsv"), row.names = FALSE, sep = "\t") #422
```

### Match biosample accession ids

```{r}
bact_attributes <- read.delim(paste0(preprocessing_path, "sra-submission/16s-attributes.tsv")) %>% #220 new IDs, but includes negative ctrls
  dplyr::select("biosample_accession" = accession, "16s_sample_id" = sample_name) %>% 
  filter(`16s_sample_id` != "16S.B0003712") %>% #remove mismatched sample
  left_join(bact_sample_ids) %>%
  dplyr::select(biosample_accession, Patient_days, Patient_days_lobe, `16s_sample_id`) %>%
  full_join(merged_attributes %>% dplyr::select(biosample_accession, Patient_days, Patient_days_lobe, `16s_sample_id`) %>% filter(!is.na(biosample_accession), !is.na(`16s_sample_id`)) )

#merge with previous IDs
merged_attributes2 <- bact_attributes %>% 
  full_join(rna_attributes) %>%
  full_join(fungi_sample_ids)

table(is.na(merged_attributes2$`16s_sample_id`)) #422
table(is.na(merged_attributes2$`its_sample_id`)) #200
table(is.na(merged_attributes2$rna_sample_id)) #234

merged_attributes2 %>% filter(!is.na(rna_sample_id)) %>% dplyr::count(is.na(biosample_accession)) #234 all have id
merged_attributes2 %>% filter(!is.na(`16s_sample_id`)) %>% dplyr::count(is.na(biosample_accession)) #422 all have id
merged_attributes2 %>% filter(!is.na(`its_sample_id`)) %>% dplyr::count(is.na(biosample_accession)) #194 with id, 6 without

table(duplicated(merged_attributes2$Patient_days_lobe))

a <- merged_attributes2 %>% filter(!is.na(Patient_days))
table(duplicated(a$Patient_days)) #403 FALSE

#merged_attributes2 %>% filter(Patient_days %in% datasets[[1]]$Patient_days) #403
```

## ITS

Use MIMARKS: specimen, human-associated

200 total + 50 ctrls (B27 samples not included)

```{r}
# Sample metadata
cols <- data.frame("organism" = "Homo sapiens ITS metagenome", 
          "isolate" = NA, 
          "collection_date" = NA,
          "env_broad_scale" = "Lower human respiratory tract",
          "env_local_scale"	= "not applicable",
          "env_medium" = "not applicable",
          "geo_loc_name" = "Australia",
          "host" = "Homo sapiens",
          "isol_growth_condt" = "not applicable",
          "lat_lon" = "not collected",
          "tissue" = "Lungs"
          )

f <- cols %>% 
  cross_join(data.frame(sample_data(fungi), row.names = NULL) %>% 
              dplyr::select(Patient_days_lobe, Record.ID, Time_interval_detailed, Group, Status) ) %>%
  left_join(data.frame(sample_data(fungi.filt), row.names = NULL) %>% 
               dplyr::select(SampleID, MiSeqID, Batch, "Patient_days_lobe" = Patient_days) ) %>% 
  left_join(patients_reassigned) %>%
  left_join(merged_attributes2 %>% dplyr::select(biosample_accession, "SampleID" = `its_sample_id`)) %>%
  dplyr::select(biosample_accession,
                "sample_name" = SampleID, 
                colnames(cols),
                "dna_extraction_id" = Batch,
                "sequencing_id" = MiSeqID, 
                "patient_id" = patient_id, 
                "disease" = Group,
                "disease_stage" = Status,
                "time_post_transplant" = Time_interval_detailed) %>%
  mutate("age" = NA, sex = NA, "sample_type" = "broncho-alveolar lavage")

# include negative and positive controls
ctrls <- cols %>%
  cross_join(data.frame(sample_data(fungi.raw))) %>%
  filter(Type != "Sample") %>%
  dplyr::select("dna_extraction_id" = Batch, "sequencing_id" = MiSeqID, "sample_name" = SampleID, "sample_type" = Type, colnames(cols)) %>%
  filter(dna_extraction_id %in% c(unique(f$dna_extraction_id), "PCR"), sequencing_id %in% unique(f$sequencing_id)) %>%
  remove_rownames()

# check overlap
length(unique(ctrls$sequencing_id)) #6
length(unique(ctrls$dna_extraction_id)) #29

length(intersect(unique(ctrls$sequencing_id), unique(f$sequencing_id))) #6 matches
length(intersect(unique(ctrls$dna_extraction_id), c(unique(f$dna_extraction_id), "PCR"))) #29 matches "B2"  "B3"  "B22" "B4"  "B5"  "B21" "B1"  "B6"  "B23" "B7"  "B12" "B20" "B11" "B8"  "B13" "B9"  "B25" "B14" "B16" "B17" PCR "B18" "B10" "B15" "B19" "B28" "B24" "B26" "B29" "B30"

f <- f %>% full_join(ctrls)

write.csv(f, paste0(preprocessing_path, "sra-submission/its-samples-metadata.csv"))

identical(sample_data(fungi)$Patient_days, all_selected_samples_collected_all_datasets %>% filter(Dataset == "Fungi") %>% pull(Patient_days))
```

```{r}
# SRA metadata
f_sra <- data.frame("sample_name" = f$sample_name,
                   "library_ID" = f$sample_name,
                   "title" = "ITS sequencing of Homo sapiens: bronchoalveolar lavage of transplanted lung",
                   "library_strategy" = "AMPLICON",
                   "library_source" = "METAGENOMIC",
                   "library_selection" = "RANDOM",
                   "library_layout" = "paired",
                   "platform" = "ILLUMINA",
                   "instrument_model" = "Illumina Miseq",
                   "design_description" = NA,
                   "filetype" = "fastq") %>%
  left_join(f) %>%
  mutate(filename = paste(sample_name, "R1.fastq.gz", sep = "-"), 
         filename2 = paste(sample_name, "R2.fastq.gz", sep = "-"))

f_sra_sub1 <- f_sra %>% filter(!is.na(biosample_accession)) %>% dplyr::select(!sample_name) #cannot have both,  194
f_sra_sub2 <- f_sra %>% filter(is.na(biosample_accession)) %>% dplyr::select(!biosample_accession) #56 = 50ctrls + 6 samples

table(duplicated(f_sra_sub1$biosample_accession))

write.table(f_sra_sub1, paste0(preprocessing_path, "sra-submission/ITS-sra-metadata-sub1.tsv"), row.names = FALSE, sep = "\t")
write.table(f_sra_sub2, paste0(preprocessing_path, "sra-submission/ITS-sra-metadata-sub2.tsv"), row.names = FALSE, sep = "\t")
write.table(f_sra, paste0(preprocessing_path, "sra-submission/ITS-sra-metadata.tsv"), row.names = FALSE, sep = "\t")
```




