---
title: "Clinical metadata"
author: "Giulia Iacono"
date: Sys.Date()
output: html_document
---

# Clinical metadata
# Datasets table

```{r}
dsa <- master_metadata$dsa_test %>% 
  filter(DSA.test == "Yes") %>%
  dplyr::select(Record.ID, DSA.Result) %>% 
  distinct() %>% 
  mutate(DSA.Result = gsub("Postitive", "Positive", DSA.Result)) %>%
  arrange(Record.ID, desc(DSA.Result) ) %>%
  group_by(Record.ID) %>%
  slice_head(., n = 1) # only one instance, negative if always negative, positive if at least 1 positive instance

pgd <- master_metadata$general_metadata %>%
  mutate(PFRatioat24hrs = as.numeric(PFRatioat24hrs),
         PGD = case_when(PFRatioat24hrs >= 300 ~ "1",
                        PFRatioat24hrs < 300 & PFRatioat24hrs > 200 ~ "2",
                        PFRatioat24hrs <= 200 ~ "3",
                        is.na(PFRatioat24hrs) ~ "Unknown")) %>%
    dplyr::select(Record.ID, PGD) 

acr <- master_metadata$biopsy %>%
  dplyr::select(Record.ID, biopsy_result) %>%
  distinct() %>% # patient 10 had 2 instances
  mutate(biopsy_result = factor(biopsy_result, level = c("A2", "A1", "A0", "Not Able to Evaluate", "Not Done"))) %>%
  arrange(Record.ID, biopsy_result ) %>%
  group_by(Record.ID) %>%
  slice_head(., n = 1) %>% # take first instance in order of factor level
  filter(Record.ID %in% unique(c(biomarker_patients, stable_patients)) ) #69 patients in total
```

```{r}
datasets <- list()

metadata_features <- c("Patient_days", "Record.ID", 
                       "Time_interval_detailed", "Time_interval_combined", "Days_interval",
  "Age", "Gender", "Donor_age", "Donor_smoking", "GIT_ischemic_time", "Group", 
  "Days", "Transplant_indication", "TypeofTransplant", 
  "DSAatTransplant", "CMVMismatch", "ICUHours", 
  "InductionTacrolimus", "InductionAzathioprine", "InductionMycophenolate", "InductionBasilimax", 
  meds_list$Name, "BAL_micro_results")

## Bacteria
b <- bact_metadata %>% 
  dplyr::select(all_of(metadata_features)) %>%
  mutate(TypeofTransplant = gsub("LS|RS", "S", TypeofTransplant)) %>%
  left_join(pgd) %>%
  left_join(dsa) %>%
  left_join(acr) %>%
  mutate(DSA.Result = ifelse(is.na(DSA.Result), "Not done", DSA.Result) ) # not done if patient was not present in dsa dataset

## Fungi
f <- fungi_metadata %>% 
  dplyr::select(all_of(metadata_features)) %>%
  mutate(TypeofTransplant = gsub("LS|RS", "S", TypeofTransplant)) %>%
  left_join(pgd) %>%
  left_join(dsa) %>%
  left_join(acr) %>%
  mutate(DSA.Result = ifelse(is.na(DSA.Result), "Not done", DSA.Result) ) # not done if patient was not present in dsa dataset

# RNA
r <- rna_corrected$samples %>% # correct has all samples
              dplyr::select(all_of(metadata_features)) %>%
  mutate(TypeofTransplant = gsub("LS|RS", "S", TypeofTransplant)) %>%
  left_join(pgd) %>%
  left_join(dsa) %>%
  left_join(acr) %>%
  mutate(DSA.Result = ifelse(is.na(DSA.Result), "Not done", DSA.Result) ) # not done if patient was not present in dsa dataset

# Molecules
m <- small_molecules_metadata %>%
  dplyr::select(all_of(metadata_features)) %>%
  mutate(TypeofTransplant = gsub("LS|RS", "S", TypeofTransplant)) %>%
  left_join(pgd) %>%
  left_join(dsa) %>%
  left_join(acr) %>%
  mutate(DSA.Result = ifelse(is.na(DSA.Result), "Not done", DSA.Result) ) # not done if patient was not present in dsa dataset()

# Total (1 always total to line up the rest of the rows)
datasets[[1]] <- b %>% full_join(f) %>% full_join(r) %>% full_join(m)
datasets[[2]] <- b 
datasets[[3]] <- f
datasets[[4]] <- r
datasets[[5]] <- m
```


```{r}
tab <- list()

for (i in c(1:length(datasets))){
  
  table <- list()

  ## Patients
  col <- unique(c(na.omit(datasets[[i]][, c("Record.ID")])))
  table[["Record.ID"]] <-  data.frame(cbind("Patients n" , paste0(length(col)), length(col))) %>% unname()

  
## Recipient Age
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Age) %>%
    distinct() %>%
    pull(Age)
  table[["Recipient_age"]] <-  data.frame(cbind("Recipient age at Tx (range)" , paste0(round(mean(col), 1), " (", paste0(min(col), "-", max(col)), ")"), round(mean(col), 1))) %>% unname()
  
  
  ## Female patients
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Gender) %>%
    distinct() %>%
    group_by(Gender) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    filter(Gender == "Female")
  table[["Gender"]] <-  data.frame(cbind("Female patients n (%)" , paste0(col$n, " (", col$Percentage, "%)"), col$n)) %>% unname()
  
  
  ## Donor Age
col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Donor_age) %>%
    distinct() %>%
    pull(Donor_age)
  table[["Donor_age"]] <-  data.frame(cbind("Donor age at Tx (range)" , paste0(round(mean(col), 1), " (", paste0(min(col), "-", max(col)), ")"), round(mean(col), 1))) %>% unname()
  
    ## Donor smoking
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Donor_smoking) %>%
    distinct() %>%
    group_by(Donor_smoking) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    filter(Donor_smoking == "TRUE")
  table[["Smoking"]] <-  data.frame(cbind("Donor smoking n (%)" , paste0(col$n, " (", col$Percentage, "%)"), col$n)) %>% unname()
 
  ## Ischemic time
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, GIT_ischemic_time) %>%
    distinct() %>%
    na.omit() %>%
    pull(GIT_ischemic_time) 
  table[["GIT_ischemic_time"]] <-  data.frame(cbind("Ischemic time (min) (range)" , paste0(round(mean(col), 1), " (", paste0(min(col), "-", max(col)), ")"), round(mean(col), 1))) %>% unname() 
  

  
## Transplant type patients
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, TypeofTransplant) %>%
    distinct() %>%
    group_by(TypeofTransplant) %>%
    dplyr::count() %>%
    rename(numb = n) %>%
    mutate(Percentage = round(numb/sum(.$numb)*100)) %>%
    rename(`Transplant type patients n (%)` = TypeofTransplant) %>%
    mutate(" " = paste0(numb, " (", Percentage, "%)")) 
  
  table[["TypeofTransplant"]] <-  rbind(gsub("numb", "", colnames(col)[c(1,4,2)]), col[, c(1,4,2)] %>% as.matrix() %>% unname())

  
## DSA at transplant patients
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, DSAatTransplant) %>%
    distinct() %>%
    group_by(DSAatTransplant) %>%
    dplyr::count() %>%
    rename(numb = n) %>%
    na.omit() %>%
    mutate(Percentage = round(numb/sum(.$numb)*100)) %>%
    filter(DSAatTransplant == "DSAatTransplant") %>%
    mutate(DSAatTransplant = gsub("DSAatTransplant", "DSA at tx patients n (%)", DSAatTransplant)) %>%
    mutate("p" = paste0(numb, " (", Percentage, "%)")) %>%
    as.matrix() %>%
    unname()

  table[["DSAatTransplant"]] <-  data.frame(cbind(col[,1], col[,4], col[,2])) %>% unname()
  
## Patients with at least one DSA after tx
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, DSA.Result) %>%
    distinct() %>%
    group_by(DSA.Result) %>%
    dplyr::count() %>%
    rename(numb = n) %>%
    mutate(Percentage = round(numb/sum(.$numb)*100)) %>%
    filter(DSA.Result == "Positive") %>%
    mutate(DSA.Result = gsub("Positive", "Positive DSA after tx patients n (%)", DSA.Result)) %>%
    mutate("p" = paste0(numb, " (", Percentage, "%)")) %>%
    as.matrix() %>%
    unname()
  
  table[["DSA.Result"]] <-  data.frame(cbind(col[,1], col[,4], col[,2])) %>% unname()

  
# ACR 
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, biopsy_result) %>%
    distinct() %>%
    group_by(biopsy_result) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    rename(`ACR after tx patients n (%)` = biopsy_result) %>%
    mutate("p" = paste0(n, " (", Percentage, "%)")) 

  table[["biopsy_result"]] <-  rbind(c(colnames(col)[c(1)], gsub("n |\\(%)|n|p", "", colnames(col)[c(4,2)])), 
                           col[, c(1,4,2)] %>% as.matrix() %>% unname())

## CMV mismatch
  col <- datasets[[1]] %>%
    dplyr::select(Record.ID, CMVMismatch) %>%
    distinct() %>%
    group_by(CMVMismatch) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    filter(CMVMismatch == "CMVMismatch") %>%
    mutate(CMVMismatch = gsub("CMVMismatch", "CMV mismatched patients n (%)", CMVMismatch)) %>%
    mutate("p" = paste0(n, " (", Percentage, "%)")) %>%
    as.matrix() %>%
    unname()

table[["CMVMismatch"]] <-  data.frame(cbind(col[,1], col[,4],col[,2])) %>% unname()

  
## ICUHours
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, ICUHours) %>%
    distinct() %>%
    pull(ICUHours)
  table[["ICUHours"]] <-  data.frame(cbind("ICU hours per patient (range)" , paste0(round(mean(col), 1), " (", paste0(min(col), "-", max(col)), ")"), round(mean(col), 1))) %>% unname()
  
## PGD grade
  ### patients
   col <- datasets[[i]] %>%
    dplyr::select(Record.ID, PGD) %>%
    distinct() %>%
    group_by(PGD) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    rename(`PGD grade patients n (%)` = PGD) %>%
    mutate(`n (%)` = paste0(n, " (", Percentage, "%)"))

  table[["PGD"]] <-  rbind(c(colnames(col)[c(1)], gsub("n |\\(%)|n", "", colnames(col)[c(4,2)])), col[, c(1,4,2)] %>% as.matrix() %>% unname())

  
  
## CLAD 
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Group) %>%
    distinct() %>%
    group_by(Group) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    filter(Group == "CLAD") %>%
    mutate(Group = gsub("CLAD", "CLAD patients n (%)", Group)) %>%
    mutate("p" = paste0(n, " (", Percentage, "%)")) %>%
    as.matrix() %>%
    unname()
  table[["Group"]] <-  data.frame(cbind("CLAD patients n (%)" , col[,4])) %>% unname()

  
## Samples
  col <- datasets[[i]][, c("Patient_days")] 
  table[["Samples"]] <-  data.frame(cbind("Samples n" , length(col),length(col))) %>% unname()
  
  ## Samples per patient
   col <- datasets[[i]] %>%
    group_by(Record.ID) %>%
    dplyr::count() 
  table[["Samples_patient"]] <-  data.frame(cbind("Samples per patient n (range)" , paste0(round(mean(col$n)), " (", paste0(min(col$n), "-", max(col$n)), ")"), round(mean(col$n)))) %>% unname()

  
## Average sampling in days
   col <- datasets[[i]] %>%
    pull(Days)
  table[["Samples_days"]] <-  data.frame(cbind("Average sampling days (range)" , paste0(round(mean(col), 1), " (", paste0(min(col), "-", max(col)), ")"), round(mean(col), 1))) %>% unname()
  
  
  ## Transplant indication
  ### patients
   col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Transplant_indication) %>%
    distinct() %>%
    group_by(Transplant_indication) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    rename(`Tx indication patients n (%)` = Transplant_indication) %>%
    mutate(`n (%)` = paste0(n, " (", Percentage, "%)"))

  table[["Transplant_indication"]] <-  rbind(c(colnames(col)[c(1)], gsub("n |\\(%)|n", "", colnames(col)[c(4,2)])), col[, c(1,4,2)] %>% as.matrix() %>% unname())

  
### Induction patients
  induction <- datasets[[i]] %>% 
  dplyr::select(Record.ID, Group, InductionTacrolimus, InductionAzathioprine, InductionBasilimax, InductionMycophenolate) %>%
    distinct()
  for (e in 3:dim(induction)[2]){induction[,e] <- ifelse(grepl("No", induction[,e]), 0, 1)}
 
induction <- induction %>%
  rowwise(Record.ID) %>%
  mutate(None = sum(InductionAzathioprine, InductionTacrolimus, InductionBasilimax, InductionMycophenolate)) 
colnames(induction) <- gsub("Induction", "", colnames(induction))

 col <- induction %>%
    pivot_longer(Tacrolimus:None, names_to = "Induction", values_to = "Count") %>%
    filter(Induction != "None" & Count != 0 | Induction == "None" & Count == 0) %>%
    group_by(Induction) %>%
    dplyr::count() %>%
    filter(Induction == "Basilimax") %>%
    rename(numb = n) %>%
    ungroup() %>%
    #arrange(desc(n)) %>%
    rename(`Induction immunosuppression patients n (%)` = Induction) %>%
    mutate(Percentage = round(numb/dim(induction)[1]*100), # corresponds to the total number of patients in that dataset
           " " = paste0(numb, " (", Percentage, "%)")) 
  table[["Induction"]] <-  rbind(gsub("numb", "", colnames(col)[c(1,4,2)]), col[, c(1,4,2)]) %>% as.matrix() %>% unname()
  

## Immunosuppressants
  immunos <- datasets[[i]] %>% 
  dplyr::select(Record.ID, Patient_days, Group, meds_list %>% filter(Type %in% c("Immunosuppressant", "Steroid")) %>% pull(Name)) %>%
    distinct()
  for (e in 4:dim(immunos)[2]){immunos[,e] <- ifelse(grepl("No", immunos[,e]), 0, 1)}
  
  ### Patients
  col <- immunos %>%
    pivot_longer(Azathioprine:Prednisolone, names_to = "Immunosuppressant", values_to = "Count") %>%
    filter(Count != 0) %>%
    group_by(Immunosuppressant, Record.ID) %>%
    dplyr::count() %>%
    group_by(Immunosuppressant) %>%
    dplyr::count() %>%
    ungroup() %>%
    #arrange(desc(n)) %>%
    #add_row(Immunosuppressant = "Other", n = sum(.$n[.$n < 9])) %>%
    #filter(n >= 9) %>%
    rename(`Maintenance immunosuppression patients n (%)` = Immunosuppressant) %>%
    mutate(Percentage = round(n/length(unique(immunos$Record.ID))*100),
           `n (%)` = paste0(n, " (", Percentage, "%)")) 
  table[["Immunosuppressant"]] <-  rbind(c(colnames(col)[c(1)], gsub("n |\\(%)|n", "", colnames(col)[c(4,2)])), col[, c(1,4,2)] %>% as.matrix() %>% unname())


## Antibiotics all
  ### Samples
  antibiotics <- datasets[[i]] %>% 
  dplyr::select(Record.ID, Patient_days, meds_list %>% filter(Type == "Antibiotic") %>% pull(Name)) %>%
    distinct()
  for (f in 3:dim(antibiotics)[2]){antibiotics[,f] <- ifelse(grepl("No", antibiotics[,f]), 0, 1)}
  
## Azithromycin alone patients
col <- antibiotics %>%
  dplyr::select(Record.ID, Azithromycin) %>%
    distinct() %>%
    group_by(Azithromycin) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    filter(Azithromycin == 1) %>%
    rename(`Azithromycin patients n (%)` = Azithromycin) %>%
    mutate(`n (%)` = paste0(n, " (", Percentage, "%)")) %>%
    as.matrix() %>%
    unname()
  table[["Azithromycin"]] <-  data.frame(cbind("Azithromycin patients n (%)" , col[,4])) %>% unname()

### Without azithromycin patients
  table[["Antibiotics"]] <- antibiotics %>%
    dplyr::select(!Azithromycin) %>%
    pivot_longer(Amoxycillin:Vancomycin, names_to = "Antibiotic", values_to = "Count") %>%
    filter(Count != 0) %>%
    group_by(Antibiotic, Record.ID) %>%
    dplyr::count() %>%
    group_by(Antibiotic) %>%
    dplyr::count() %>%
    ungroup()
  
    if (i == 1){ # based on the total column to line up the rows, always 1
    ant.other <- table[["Antibiotics"]] %>% filter(n < 5) %>% pull(Antibiotic) 
    }
  
  col <- table[["Antibiotics"]] %>% 
    add_row(`Antibiotic` = "Other", n = sum(.$n[.$Antibiotic %in% ant.other])) %>%
    filter(!Antibiotic %in% ant.other) %>%
    rename(`Antibiotics patients n (%)` = Antibiotic) %>%
    mutate(Percentage_pat = round(n/length(unique(antibiotics$Record.ID))*100),
           " " = paste0(`n`, " (", Percentage_pat, "%)")) %>% 
    dplyr::select(1,4,2) # 3 correct n_pat
  
    table[["Antibiotics"]] <-  rbind(c(colnames(col)[c(1)], gsub("n", "", colnames(col)[c(2,3)])), col %>% as.matrix() %>% unname())
  
  
## Positive microbiology
  ### patients
   col <- datasets[[i]] %>%
    dplyr::select(Record.ID, BAL_micro_results) %>%
    na.omit() %>%
    distinct() %>%
    group_by(BAL_micro_results) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    rename(`BAL microbiology patients n (%)` = BAL_micro_results) %>%
    mutate(`n (%)` = paste0(n, " (", Percentage, "%)"))
  
  table[["BAL_micro_results"]] <- rbind(c(colnames(col)[c(1)], gsub("n |\\(%)|n", "", colnames(col)[c(4,2)])), col[, c(1,4,2)] %>% as.matrix() %>% unname())
  
tab[[i]] <- plyr::ldply(table, data.frame)    
  
}

datasets_full_table <- tab[[1]] %>% dplyr::select("ID" = `.id`, "demo" = X1, "Total" = X2) %>%
  full_join(tab[[2]] %>% dplyr::select("ID" = `.id`, "demo" = X1, "16S Bacteria" = X2)) %>%
  full_join(tab[[3]] %>% dplyr::select("ID" = `.id`, "demo" = X1, "ITS Fungi" = X2)) %>%
  full_join(tab[[4]] %>% dplyr::select("ID" = `.id`, "demo" = X1, "RNA" = X2)) %>%
  full_join(tab[[5]] %>% dplyr::select("ID" = `.id`, "demo" = X1, "Small molecules" = X2)) %>%
  mutate(`Small molecules` = ifelse(is.na(`Small molecules`), "0 (0%)", `Small molecules`))
write.csv(datasets_full_table, paste0(analysis_path, "clinical-metadata/tables/datasets.csv"), row.names = FALSE)

rm(col)
```


# Biomarker cohort table

```{r}
tab <- list()

# Use total only
datasets[[1]] <- datasets[[1]] %>% filter(Record.ID %in% biomarker_patients) 
datasets[[2]] <- datasets[[1]] %>% filter(Group == "CLAD-free" & Record.ID %in% biomarker_patients) 
datasets[[3]] <- datasets[[1]] %>% filter(Group == "CLAD" & Record.ID %in% biomarker_patients) 
# leave 4 empty

for (i in c(1:length(datasets))[-c(4, 5)]){
  
  table <- list()

  ## Patients
  col <- unique(c(na.omit(datasets[[i]][, c("Record.ID")])))
  table[["Record.ID"]] <-  data.frame(cbind("Patients n" , paste0(length(col)), length(col))) %>% unname()

  
## Recipient age
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Age) %>%
    distinct() %>%
    pull(Age)
  table[["Age"]] <-  data.frame(cbind("Recipient age at Tx (range)" , paste0(round(mean(col), 1), " (", paste0(min(col), "-", max(col)), ")"), round(mean(col), 1))) %>% unname()
  
  
  ## Female patients
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Gender) %>%
    distinct() %>%
    group_by(Gender) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    filter(Gender == "Female")
  table[["Gender"]] <-  data.frame(cbind("Female patients n (%)" , paste0(col$n, " (", col$Percentage, "%)"), col$n)) %>% unname()
  
    ## Donor Age
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Donor_age) %>%
    distinct() %>%
    pull(Donor_age)
  table[["Donor_age"]] <-  data.frame(cbind("Donor age at Tx (range)" , paste0(round(mean(col), 1), " (", paste0(min(col), "-", max(col)), ")"), round(mean(col), 1))) %>% unname()
  
    ## Donor smoking
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Donor_smoking) %>%
    distinct() %>%
    group_by(Donor_smoking) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    filter(Donor_smoking == "TRUE")
  table[["Donor_smoking"]] <-  data.frame(cbind("Donor smoking n (%)" , paste0(col$n, " (", col$Percentage, "%)"), col$n)) %>% unname()
 
  ## Ischemic time
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, GIT_ischemic_time) %>%
    distinct() %>%
    na.omit() %>%
    pull(GIT_ischemic_time) 
  table[["GIT_ischemic_time"]] <-  data.frame(cbind("Ischemic time (min) (range)" , paste0(round(mean(col), 1), " (", paste0(min(col), "-", max(col)), ")"), round(mean(col), 1))) %>% unname() 

  
## Transplant type patients
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, TypeofTransplant) %>%
    distinct() %>%
    group_by(TypeofTransplant) %>%
    dplyr::count() %>%
    rename(numb = n) %>%
    mutate(Percentage = round(numb/sum(.$numb)*100)) %>%
    rename(`Transplant type patients n (%)` = TypeofTransplant) %>%
    mutate(" " = paste0(numb, " (", Percentage, "%)")) 
  
  table[["TypeofTransplant"]] <-  rbind(gsub("numb", "", colnames(col)[c(1,4,2)]), col[, c(1,4,2)] %>% as.matrix() %>% unname())

  
## DSA at transplant patients
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, DSAatTransplant) %>%
    distinct() %>%
    group_by(DSAatTransplant) %>%
    dplyr::count() %>%
    rename(numb = n) %>%
    na.omit() %>%
    mutate(Percentage = round(numb/sum(.$numb)*100)) %>%
    filter(DSAatTransplant == "DSAatTransplant") %>%
    mutate(DSAatTransplant = gsub("DSAatTransplant", "DSA at tx patients n (%)", DSAatTransplant)) %>%
    mutate("p" = paste0(numb, " (", Percentage, "%)")) %>%
    as.matrix() %>%
    unname()

  table[["DSAatTransplant"]] <-  data.frame(cbind(col[,1], col[,4], col[,2])) %>% unname()
  

  ## Patients with at least one DSA after tx
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, DSA.Result) %>%
    distinct() %>%
    group_by(DSA.Result) %>%
    dplyr::count() %>%
    rename(numb = n) %>%
    mutate(Percentage = round(numb/sum(.$numb)*100)) %>%
    filter(DSA.Result == "Positive") %>%
    mutate(DSA.Result = gsub("Positive", "Positive DSA after tx patients n (%)", DSA.Result)) %>%
    mutate("p" = paste0(numb, " (", Percentage, "%)")) %>%
    as.matrix() %>%
    unname()
  
  table[["DSA.Result"]] <-  data.frame(cbind(col[,1], col[,4], col[,2])) %>% unname()
  
  
# ACR 
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, biopsy_result) %>%
    distinct() %>%
    group_by(biopsy_result) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    rename(`ACR after tx patients n (%)` = biopsy_result) %>%
    mutate("p" = paste0(n, " (", Percentage, "%)")) 

  table[["biopsy_result"]] <-  rbind(c(colnames(col)[c(1)], gsub("n |\\(%)|n|p", "", colnames(col)[c(4,2)])), 
                           col[, c(1,4,2)] %>% as.matrix() %>% unname())

## CMV mismatch
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, CMVMismatch) %>%
    distinct() %>%
    group_by(CMVMismatch) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    filter(CMVMismatch == "CMVMismatch") %>%
    mutate(CMVMismatch = gsub("CMVMismatch", "CMV mismatched patients n (%)", CMVMismatch)) %>%
    mutate("p" = paste0(n, " (", Percentage, "%)")) %>%
    as.matrix() %>%
    unname()

table[["CMVMismatch"]] <-  data.frame(cbind(col[,1], col[,4],col[,2])) %>% unname()


## ICUHours
  col <- datasets[[i]] %>%
    dplyr::select(Record.ID, ICUHours) %>%
    distinct() %>%
    pull(ICUHours)
  table[["ICUHours"]] <-  data.frame(cbind("ICU hours per patient (range)" , paste0(round(mean(col), 1), " (", paste0(min(col), "-", max(col)), ")"), round(mean(col), 1))) %>% unname()
  

## PGD grade
  ### patients
   col <- datasets[[i]] %>%
    dplyr::select(Record.ID, PGD) %>%
    distinct() %>%
    group_by(PGD) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    rename(`PGD grade patients n (%)` = PGD) %>%
    mutate(`n (%)` = paste0(n, " (", Percentage, "%)"))

  table[["PGD"]] <-  rbind(c(colnames(col)[c(1)], gsub("n |\\(%)|n", "", colnames(col)[c(4,2)])), col[, c(1,4,2)] %>% as.matrix() %>% unname())

  
## Samples
  col <- datasets[[i]][, c("Patient_days")] 
  table[["Samples"]] <-  data.frame(cbind("Samples n" , length(col),length(col))) %>% unname()
  
  
  ## Samples per patient
   col <- datasets[[i]] %>%
    group_by(Record.ID) %>%
    dplyr::count() 
  table[["Samples_patient"]] <-  data.frame(cbind("Samples per patient (range)" , paste0(round(mean(col$n)), " (", paste0(min(col$n), "-", max(col$n)), ")"), round(mean(col$n)))) %>% unname()

  
## Average sampling in days
   col <- datasets[[i]] %>%
    pull(Days)
  table[["Days"]] <-  data.frame(cbind("Average sampling days (range)" , paste0(round(mean(col), 1), " (", paste0(min(col), "-", max(col)), ")"), round(mean(col), 1))) %>% unname()
  
  

 ## Transplant indication
  ### patients
   col <- datasets[[i]] %>%
    dplyr::select(Record.ID, Transplant_indication) %>%
    distinct() %>%
    group_by(Transplant_indication) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    rename(`Tx indication patients n (%)` = Transplant_indication) %>%
    mutate(`n (%)` = paste0(n, " (", Percentage, "%)"))

  table[["Transplant_indication"]] <-  rbind(c(colnames(col)[c(1)], gsub("n |\\(%)|n", "", colnames(col)[c(4,2)])), col[, c(1,4,2)] %>% as.matrix() %>% unname())


### Induction patients
  induction <- datasets[[i]] %>% 
  dplyr::select(Record.ID, Group, InductionTacrolimus, InductionAzathioprine, InductionBasilimax, InductionMycophenolate) %>%
    distinct()
  for (e in 3:dim(induction)[2]){induction[,e] <- ifelse(grepl("No", induction[,e]), 0, 1)}
 
induction <- induction %>%
  rowwise(Record.ID) %>%
  mutate(None = sum(InductionAzathioprine, InductionTacrolimus, InductionBasilimax, InductionMycophenolate)) 
colnames(induction) <- gsub("Induction", "", colnames(induction))

 col <- induction %>%
    pivot_longer(Tacrolimus:None, names_to = "Induction", values_to = "Count") %>%
    filter(Induction != "None" & Count != 0 | Induction == "None" & Count == 0) %>%
    group_by(Induction) %>%
    dplyr::count() %>%
    filter(Induction == "Basilimax") %>%
    rename(numb = n) %>%
    ungroup() %>%
    #arrange(desc(n)) %>%
    rename(`Induction immunosuppression patients n (%)` = Induction) %>%
    mutate(Percentage = round(numb/dim(induction)[1]*100), # corresponds to the total number of patients in that dataset
           " " = paste0(numb, " (", Percentage, "%)")) 
  table[["Induction"]] <-  rbind(gsub("numb", "", colnames(col)[c(1,4,2)]), col[, c(1,4,2)]) %>% as.matrix() %>% unname()
  
  
  
## Immunosuppressants
  immunos <- datasets[[i]] %>% 
  dplyr::select(Record.ID, Patient_days, Group, meds_list %>% filter(Type %in% c("Immunosuppressant", "Steroid")) %>% pull(Name)) %>%
    distinct()
  for (e in 4:dim(immunos)[2]){immunos[,e] <- ifelse(grepl("No", immunos[,e]), 0, 1)}
  
  ### Patients
  col <- immunos %>%
    pivot_longer(Azathioprine:Prednisolone, names_to = "Immunosuppressant", values_to = "Count") %>%
    filter(Count != 0) %>%
    group_by(Immunosuppressant, Record.ID) %>%
    dplyr::count() %>%
    group_by(Immunosuppressant) %>%
    dplyr::count() %>%
    ungroup() %>%
    #arrange(desc(n)) %>%
    #add_row(Immunosuppressant = "Other", n = sum(.$n[.$n < 9])) %>%
    #filter(n >= 9) %>%
    rename(`Maintenance immunosuppression patients n (%)` = Immunosuppressant) %>%
    mutate(Percentage = round(n/length(unique(immunos$Record.ID))*100),
           `n (%)` = paste0(n, " (", Percentage, "%)")) 
  table[["Immunosuppressant"]] <-  rbind(c(colnames(col)[c(1)], gsub("n |\\(%)|n", "", colnames(col)[c(4,2)])), col[, c(1,4,2)] %>% as.matrix() %>% unname())
  
  
## Antibiotics
  ### Samples
  antibiotics <- datasets[[i]] %>% 
  dplyr::select(Record.ID, Patient_days, Group, meds_list %>% filter(Type == "Antibiotic") %>% pull(Name)) %>%
    distinct()
  for (f in 4:dim(antibiotics)[2]){antibiotics[,f] <- ifelse(grepl("No", antibiotics[,f]), 0, 1)}
  
  ## Azithromycin alone patients
col <- antibiotics %>%
  dplyr::select(Record.ID, Azithromycin) %>%
    distinct() %>%
    group_by(Azithromycin) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    filter(Azithromycin == 1) %>%
    rename(`Azithromycin patients n (%)` = Azithromycin) %>%
    mutate(`n (%)` = paste0(n, " (", Percentage, "%)")) %>%
    as.matrix() %>%
    unname()
  table[["Azithromycin"]] <-  data.frame(cbind("Azithromycin patients n (%)" , col[,c(4)])) %>% unname()
  
  ### Without azithromycin patients
  table[["Antibiotics"]] <- antibiotics %>%
    dplyr::select(!Azithromycin) %>%
    pivot_longer(Amoxycillin:Vancomycin, names_to = "Antibiotic", values_to = "Count") %>%
    filter(Count != 0) %>%
    group_by(Antibiotic, Record.ID) %>%
    dplyr::count() %>%
    group_by(Antibiotic) %>%
    dplyr::count() %>%
    ungroup()
  
    if (i == 1){ # based on the total column to line up the rows, always 1
    ant.other <- table[["Antibiotics"]] %>% filter(n < 3) %>% pull(Antibiotic) 
    }
  
  col <- table[["Antibiotics"]] %>% 
    add_row(`Antibiotic` = "Other", n = sum(.$n[.$Antibiotic %in% ant.other])) %>%
    filter(!Antibiotic %in% ant.other) %>%
    rename(`Antibiotics patients n (%)` = Antibiotic) %>%
    mutate(Percentage_pat = round(n/length(unique(antibiotics$Record.ID))*100),
           " " = paste0(`n`, " (", Percentage_pat, "%)")) %>% 
    dplyr::select(1,4,2) # 3 correct n_pat
  
    table[["Antibiotics"]] <-  rbind(c(colnames(col)[c(1)], gsub("n", "", colnames(col)[c(2,3)])), col %>% as.matrix() %>% unname())

    
## Positive microbiology
  ### patients
   col <- datasets[[i]] %>%
    dplyr::select(Record.ID, BAL_micro_results) %>%
    distinct() %>%
    na.omit() %>%
    group_by(BAL_micro_results) %>%
    dplyr::count() %>%
    mutate(Percentage = round(n/sum(.$n)*100)) %>%
    rename(`BAL microbiology patients n (%)` = BAL_micro_results) %>%
    mutate(`n (%)` = paste0(n, " (", Percentage, "%)"))
  
  table[["BAL_micro_results"]] <- rbind(c(colnames(col)[c(1)], gsub("n |\\(%)|n", "", colnames(col)[c(4,2)])), col[, c(1,4,2)] %>% as.matrix() %>% unname())
  
tab[[i]] <- plyr::ldply(table, data.frame)  


## pvalue column using dataset[[1]] 
if (i == 1){
  
## Only on patients, each row is a single patient, one value per patient, no Days
  table <- list()
  c_columns <- c("Age", "Gender", "TypeofTransplant", "Donor_age", "Donor_smoking", "GIT_ischemic_time",
                                                "DSAatTransplant", "DSA.Result", "CMVMismatch",
                                                "ICUHours", "PGD", "Transplant_indication", "biopsy_result", 
                                                "Group") # leave group last
  
demographics <- datasets[[i]][, c_columns] %>% distinct() %>% group_by(Group) 
total.clad.free <- demographics %>% dplyr::count(Group) %>% filter(Group == "CLAD-free") %>% pull(n)
total.clad <- demographics %>% dplyr::count(Group) %>% filter(Group == "CLAD") %>% pull(n) #13

for (e in 1:(length(c_columns)-1)){ # -1 to leave out the Group column
  
  # If character
  if (is.character(demographics %>% pull(c_columns[e])) | is.factor(demographics %>% pull(c_columns[e])) | is.logical(demographics %>% pull(c_columns[e]))){
    
    d <- demographics[, c("Group", c_columns[e])] %>%
      dplyr::count(.[,2]) %>%
      pivot_wider(., names_from = Group, values_from = n) %>%
      na.omit() %>%
      column_to_rownames(., var = c_columns[e]) %>%
      relocate(., c("CLAD")) 
    
table[[c_columns[e]]] <- round(chisq.test(d, correct = FALSE)$p.value, 4) }
  
  # If numeric
  if (is.numeric(demographics %>% pull(c_columns[e]))){
    
  ## Perform t-test or wilcox test
  clad.test <- ungroup(demographics[, c("Group", c_columns[e])])
  
  if (shapiro.test(as.numeric(clad.test %>% pull(2)))$p.value < 0.05){
      table[[c_columns[e]]] <- round(wilcox.test(clad.test[[c_columns[e]]] ~ clad.test$Group)$p.value, 4) }
  
  if (shapiro.test(as.numeric(clad.test %>% pull(2)))$p.value >= 0.05){
      table[[c_columns[e]]] <- round(t.test(clad.test[[c_columns[e]]] ~ clad.test$Group)$p.value, 4) } 
  }
  
  
## Add induction patients
d <- induction %>%
  dplyr::select(Record.ID, Group, Basilimax) %>%
  group_by(Group) %>%
  pivot_longer(Basilimax, names_to = "Basiliximab", values_to = "Count") %>%
  group_by(Group, Count) %>%
  dplyr::count()  %>%
  pivot_wider(names_from = Group, values_from = n) %>%
  column_to_rownames("Count")
table[["Induction"]] <- round(chisq.test(d, correct = FALSE)$p.value, 4)

## Add immunos patients
d <- immunos %>%
    pivot_longer(Azathioprine:Prednisolone, names_to = "Immunosuppressant", values_to = "Count") %>%
    filter(Count != 0) %>%
    group_by(Immunosuppressant, Group, Record.ID) %>% # how many samples per patient
    dplyr::count() %>%
    filter(n != 0) %>%
    group_by(Immunosuppressant, Group) %>% # how many patients
    dplyr::count() %>%
    pivot_wider(names_from = Group, values_from = n) %>%
    column_to_rownames("Immunosuppressant") %>%
    replace(is.na(.), 0)
table[["Immunosuppressant"]] <- round(chisq.test(d, correct = FALSE)$p.value, 4)


## Antibiotics patients no azithromycin
d <- antibiotics %>%
    dplyr::select(!Azithromycin) %>%
    pivot_longer(Amoxycillin:Vancomycin, names_to = "Antibiotic", values_to = "Count") %>%
    filter(Count != 0) %>%
    group_by(Antibiotic, Group, Record.ID) %>%
    dplyr::count() %>%
    filter(n != 0) %>%
    group_by(Antibiotic, Group) %>%
    dplyr::count() %>%
    ungroup() %>%
    pivot_wider(names_from = Group, values_from = n) %>%
    replace(is.na(.), 0) %>%
    add_row(`Antibiotic` = "Other", `CLAD-free` = sum(.$`CLAD-free`[.$Antibiotic %in% ant.other]), CLAD = sum(.$CLAD[.$Antibiotic %in% ant.other])) %>%
    filter(!Antibiotic %in% ant.other) %>%
    column_to_rownames("Antibiotic") %>%
    replace(is.na(.), 0)
table[["Antibiotics"]] <- round(chisq.test(d, correct = FALSE)$p.value, 4)

d <- antibiotics %>%
  dplyr::select(Record.ID, Group, Azithromycin) %>%
  distinct() %>%
  group_by(Record.ID) %>%
  arrange(desc(Azithromycin)) %>%
  slice_head(n = 1) %>% # if they ever took it it will be equal to 1, otherwise 0
  group_by(Group) %>%
  pivot_longer(Azithromycin, names_to = "Azithromycin", values_to = "Count") %>%
  group_by(Group, Count) %>%
  dplyr::count()  %>%
  pivot_wider(names_from = Group, values_from = n) %>%
  column_to_rownames("Count")
table[["Azithromycin"]] <- round(chisq.test(d, correct = FALSE)$p.value, 4)

## Microbiology
d <- datasets[[i]] %>%
    dplyr::select(Record.ID, Group, BAL_micro_results) %>%
    distinct() %>%
    na.omit() %>%
    group_by(BAL_micro_results, Group, Record.ID) %>%
    dplyr::count() %>%
    group_by(BAL_micro_results, Group) %>%
    dplyr::count() %>%
    ungroup() %>%
    pivot_wider(names_from = Group, values_from = n) %>%
    na.omit() %>%
    column_to_rownames("BAL_micro_results")
table[["BAL_micro_results"]] <- round(chisq.test(d, correct = FALSE)$p.value, 4)

## On samples
## Add samples per patient
d <- datasets[[i]] %>%
  group_by(Group, Record.ID) %>%
  dplyr::count() %>%
  ungroup()
table[["Samples_patient"]] <-  round(t.test(d$n ~ d$Group)$p.value, 4)

## Add average sampling days per group
d <- datasets[[i]]

  if (shapiro.test(as.numeric(d %>% pull(Days)))$p.value < 0.05){
      table[["Days"]] <- round(wilcox.test(d$Days ~ d$Group)$p.value, 4) }
  
  if (shapiro.test(as.numeric(d %>% pull(Days)))$p.value >= 0.05){
      table[["Days"]] <- round(t.test(d$Days ~ d$Group)$p.value, 4) } 


if (i == 1){
  tab[[4]] <- plyr::ldply(table, data.frame) }

}

}
}

biomarker_full_table <- tab[[1]] %>% dplyr::select("ID" = `.id`, "demo" = X1, "Total" = X2) %>%
  full_join(tab[[2]] %>% dplyr::select("ID" = `.id`, "demo" = X1, "CLAD-free" = X2)) %>%
  full_join(tab[[3]] %>% dplyr::select("ID" = `.id`, "demo" = X1, "CLAD" = X2)) %>%
  full_join(tab[[4]] %>% dplyr::select("ID" = `.id`, "pvalue" = X..i..) ) %>%
  mutate(CLAD = ifelse(is.na(CLAD), "0 (%)", CLAD),
         `CLAD-free` = ifelse(is.na(`CLAD-free`), "0 (%)", `CLAD-free`))
write.csv(biomarker_full_table, paste0(analysis_path, "clinical-metadata/tables/group-demographics.csv"), row.names = FALSE)

#rm(col, induction, antibiotics, c_columns, clad.test, demographics, tab, table)
```

## DSA testing

```{r}
dsa_bronchs <- master_metadata$dsa_test %>% 
  filter(DSA.test == "Yes") %>% # no days for when not tested
  mutate(DSA.Result = gsub("Postitive", "Positive", DSA.Result)) %>%
  arrange(Record.ID, desc(DSA.Result) ) %>%
  filter(Record.ID %in% biomarker_patients) %>% 
  inner_join(clad_assessment, by = "Record.ID") %>%
  distinct() %>%
  mutate(Group = factor(Group, levels = c("CLAD-free", "CLAD")),
         days_to_clad = clad_day_clinical - Days,
         Status = ifelse(days_to_clad > 0 | is.na(days_to_clad), "(before)", "(after)"),
         CLAD_diagnosis = "CLAD diagnosis",
         Months = round(Days/30.417, 1),
         Record.ID = factor(Record.ID)) %>%
  filter(is.na(days_to_clad) | days_to_clad >= -30, Months < 26) 

p <- 
  
ggplot(dsa_bronchs, aes(x = Months, y = Record.ID)) +
  geom_path(aes(x = Months, y = as.character(Record.ID), group = Record.ID), colour = "grey60", linewidth = 0.8) + # connect bals
  geom_point(aes(fill = DSA.Result), shape = 21, size = 2) +
  geom_point(data = biopsies %>% filter(Group == "CLAD"), aes(x = round(clad_day_clinical/30.417, 1), y = as.character(Record.ID), colour = CLAD_diagnosis), shape = 4,  size = 1) + #clad day
  theme +
  theme(axis.text.y = element_blank(), legend.position = "bottom") +
  labs(title = "DSA testing", x = "Months", y = "Patient", fill = "Result", colour = "") +
  facet_grid(Group~., scales = "free", space = "free", switch = "y") +
  scale_fill_manual(values = c("Negative" = "darkgrey", "Positive" = "#8837D5")) +
  scale_colour_manual(values = c("CLAD diagnosis" = "red"))

ggsave(paste0(analysis_path, "clinical-metadata/dsa-testing.png"), p, width = 8, height = 11, units = 'cm', dpi = 600)
```

## CMV testing
### BAL

```{r}
cmv_bal <- master_metadata$bal_class_hosp %>% 
  dplyr::select(Record.ID, Days, CMVinBAL, IfPositiveCMVBALViralLoad) %>%
  filter(Record.ID %in% biomarker_patients) %>% 
  inner_join(clad_assessment, by = "Record.ID") %>%
  distinct() %>%
  mutate(Group = factor(Group, levels = c("CLAD-free", "CLAD")),
         days_to_clad = clad_day_clinical - Days,
         Status = ifelse(days_to_clad > 0 | is.na(days_to_clad), "(before)", "(after)"),
         CLAD_diagnosis = "CLAD diagnosis",
         Months = round(Days/30.417, 1),
         Record.ID = factor(Record.ID)) %>%
  filter(is.na(days_to_clad) | days_to_clad >= -30, Months < 26) %>%
  mutate(cmv_result = case_when(CMVinBAL == "CMVinBAL" ~ "Detected",
                                CMVinBAL == "NoCMVinBAL" ~ "Not Detected",
                                CMVinBAL == "Not Done" ~ "Not Done") )

p <- 
  
ggplot(cmv_bal, aes(x = Months, y = Record.ID)) +
  geom_path(aes(x = Months, y = as.character(Record.ID), group = Record.ID), colour = "grey60", linewidth = 0.8) + # connect bals
  geom_point(aes(fill = IfPositiveCMVBALViralLoad, shape = cmv_result), size = 2) +
  geom_point(data = biopsies %>% filter(Group == "CLAD"), aes(x = round(clad_day_clinical/30.417, 1), y = as.character(Record.ID), colour = CLAD_diagnosis), shape = 4,  size = 1) + #clad day
  theme +
  theme(axis.text.y = element_blank()) +
  guides(colour = "none") +
  labs(title = "CMV in BAL", x = "Months", y = "Patient", fill = "Viral load", colour = "", shape = "Result") +
  facet_grid(Group~., scales = "free", space = "free", switch = "y") +
  scale_fill_gradient(low = "grey", high = "#8837D5") +
  scale_shape_manual(values = c("Not Detected" = 21, "Detected" = 22, "Not Done" = 23)) +
  scale_colour_manual(values = c("CLAD diagnosis" = "red"))

ggsave(paste0(analysis_path, "clinical-metadata/cmv-bal.png"), p, width = 14, height = 10, units = 'cm', dpi = 600)
```

### Blood

```{r}
cmv_blood <- master_metadata$cmv_blood %>% 
  filter(Record.ID %in% biomarker_patients) %>% 
  left_join(clad_assessment, by = "Record.ID") %>%
  distinct() %>%
  mutate(Group = factor(Group, levels = c("CLAD-free", "CLAD")),
         days_to_clad = clad_day_clinical - Days,
         Status = ifelse(days_to_clad > 0 | is.na(days_to_clad), "(before)", "(after)"),
         CLAD_diagnosis = "CLAD diagnosis",
         Months = round(Days/30.417, 1),
         Record.ID = factor(Record.ID)) %>%
  filter(is.na(days_to_clad) | days_to_clad >= -30, Months < 26) %>%
  mutate(Blood_result = ifelse(CMV_blood_test == "No", "Not done", Blood_result),
         Blood_result = factor(Blood_result, levels = c("Detected", "Not Detected")))

p <- 
  
ggplot(cmv_blood, aes(x = Months, y = Record.ID)) +
  geom_path(aes(x = Months, y = as.character(Record.ID), group = Record.ID), colour = "grey60", linewidth = 0.8) + # connect bals
  geom_point(aes(fill = Viral_load, shape = Blood_result), size = 2) +
  geom_point(data = biopsies %>% filter(Group == "CLAD"), aes(x = round(clad_day_clinical/30.417, 1), y = as.character(Record.ID), colour = CLAD_diagnosis), shape = 4,  size = 1) + #clad day
  theme +
  theme(axis.text.y = element_blank()) +
  guides(colour = "none") +
  labs(title = "CMV in blood", x = "Months", y = "Patient", fill = "Viral load", colour = "", shape = "Result") +
  facet_grid(Group~., scales = "free", space = "free", switch = "y") +
  scale_colour_manual(values = c("CLAD diagnosis" = "red")) +
  scale_shape_manual(values = c("Not Detected" = 21, "Detected" = 22)) +
  scale_fill_gradient(low = "grey", high = "orange")

ggsave(paste0(analysis_path, "clinical-metadata/cmv-blood.png"), p, width = 14, height = 10, units = 'cm', dpi = 600)
```


## Acute cellular rejection

```{r}
biopsies <- master_metadata$biopsy %>%
  filter(Record.ID %in% biomarker_patients) %>% 
  inner_join(clad_assessment, by = "Record.ID") %>%
  distinct() %>%
  mutate(biopsy_result = ifelse(is.na(biopsy_result), "Not Done", biopsy_result),
         Group = factor(Group, levels = c("CLAD-free", "CLAD")),
         days_to_clad = clad_day_clinical - Days,
         Status = ifelse(days_to_clad > 0 | is.na(days_to_clad), "(before)", "(after)"),
         CLAD_diagnosis = "CLAD diagnosis",
         Months = round(Days/30.417, 1),
         Record.ID = factor(Record.ID)) %>%
  filter(is.na(days_to_clad) | days_to_clad >= -30, Months < 26) 

p <- 
  
ggplot(biopsies, aes(x = Months, y = Record.ID)) +
  geom_path(aes(x = Months, y = as.character(Record.ID), group = Record.ID), colour = "grey60", linewidth = 0.8) + # connect bals
  geom_point(aes(fill = biopsy_result), shape = 21, size = 2) +
  geom_point(data = biopsies %>% filter(Group == "CLAD"), aes(x = round(clad_day_clinical/30.417, 1), y = as.character(Record.ID), colour = CLAD_diagnosis), shape = 4,  size = 1) + #clad day
  theme +
  theme(axis.text.y = element_blank(), legend.position = "bottom") +
  guides(colour = "none") +
  labs(title = "Acute cellular rejection", x = "Months", y = "Patient", fill = "Result", colour = "") +
  facet_grid(Group~., scales = "free", space = "free", switch = "y") +
  scale_fill_manual(values = c("A0" = "darkgrey", "A1" = "#8837D5", "A2" = "#F26DF9", "Not Able to Evaluate" = "black", "Not Done" = "white")) +
  scale_colour_manual(values = c("CLAD diagnosis" = "red"))

ggsave(paste0(analysis_path, "clinical-metadata/acr.png"), p, width = 13, height = 11, units = 'cm', dpi = 600)
```

## Azithromycin

```{r}
azithromycin <- master_metadata$bronch_metadata %>% 
  filter(Record.ID %in% unique(c(biomarker_patients, stable_patients))) %>% 
  dplyr::select(Patient_days, Record.ID, Days, Azithromycin) %>%
  inner_join(clad_assessment, by = "Record.ID") %>%
  right_join(all_selected_samples_collected_across_datasets[, c("Record.ID", "Study")] %>% distinct()) %>%
  mutate(Record.ID = as.factor(Record.ID),
         Months = round(Days/30.417, 1),
         days_to_clad = clad_day_clinical - Days,
         CLAD_diagnosis = "CLAD diagnosis",
         Group = ifelse(clad_clinical == TRUE, "CLAD", "CLAD-free"),
         Group = factor(Group, levels = c("CLAD-free", "CLAD"))) %>%
  filter(is.na(days_to_clad) | days_to_clad >= -30, Months < 26) 



p <-  
  ggplot(azithromycin, aes(x = Months, y = Record.ID)) +
  geom_path(aes(group = Record.ID)) +
  geom_point(aes(fill = Azithromycin), shape = 21) +
  geom_point(data = d %>% filter(Group == "CLAD"), aes(x = round(clad_day_clinical/30.417, 1), y = Record.ID, colour = CLAD_diagnosis), shape = 4,  size = 1) + #clad day
  theme +
  theme(axis.text.y = element_blank()) +
  guides(colour = "none") +
  labs(title = "Azithromycin", y = "Patient", colour = "CLAD diagnosis") +
  facet_grid(Study+Group~., scales = "free", space = "free", switch = "y") + #
  scale_fill_manual(values = c("NoAzithromycin" = "grey", "Azithromycin" = "blue")) +
  scale_x_continuous(breaks = c(seq(from = 0, to = 26, by = 1)))

ggsave(paste0(analysis_path, "clinical-metadata/azithromycin2.png"), p, width = 13, height = 10, units = 'cm', dpi = 600)

# when was azithromycin started
azithromycin %>%
  filter(Days < 30, Azithromycin == "Azithromycin") %>%
  pull(Record.ID) %>%
  unique() %>%
  length() # 17


  ggplot(azithromycin, aes(x = Days, y = Record.ID)) +
  geom_path(aes(group = Record.ID)) +
  geom_point(aes(fill = Azithromycin), shape = 21) +
  geom_point(data = d %>% filter(Group == "CLAD"), aes(x = clad_day_clinical, y = Record.ID, colour = CLAD_diagnosis), shape = 4,  size = 1) + #clad day
  theme +
  theme(axis.text.y = element_blank()) +
  guides(colour = "none") +
  labs(title = "Azithromycin", y = "Patient", colour = "CLAD diagnosis") +
  facet_grid(Study+Group~., scales = "free", space = "free", switch = "y") + #
  scale_fill_manual(values = c("NoAzithromycin" = "grey", "Azithromycin" = "blue"))
```


### Medications

"Posaconazole", "Voriconazole", "AmphothericinB", "Ciprofloxacin", "Piperacillin-Tazobactam", "Augmentin", "Meropenem", "Tobramycin", "Trimethoprim", "Pentamidine", "Moxifloxacin", "Colistin", "Resprim", "Vancomycin", "Amoxycillin", "Isozianid", "Ceftazidime", "Flucloxacillin", , , ,  , , "BactrimDS", "Valganciclovir", "Ganciclovir", "Azithromycin",  "Valaciclovir"

```{r}
meds <- master_metadata$bronch_metadata %>%
  filter(Record.ID %in% unique(c(biomarker_patients, stable_patients))) %>% 
  dplyr::select(Record.ID, Days, "Tacrolimus", "Cyclosporine" = "Cyclosporin", "Prednisolone", "Mycophenolate-Mofetil", "Azathioprine", "Everolimus") %>%
  mutate(Cyclosporine = gsub("Cyclosporin", "Cyclosporine", Cyclosporine)) %>%
  pivot_longer(Tacrolimus:Everolimus, names_to = "Immunosuppression", values_to = "Value") %>%
  inner_join(clad_assessment, by = "Record.ID") %>%
  right_join(all_selected_samples_collected_across_datasets[, c("Record.ID", "Study")] %>% distinct()) %>%
  mutate(Record.ID = as.factor(Record.ID),
         Months = round(Days/30.417, 1),
         days_to_clad = clad_day_clinical - Days,
         CLAD_diagnosis = "CLAD diagnosis",
         Group = ifelse(clad_clinical == TRUE, "CLAD", "CLAD-free"),
         Group = factor(Group, levels = c("CLAD-free", "CLAD"))) %>%
  filter(is.na(days_to_clad) | days_to_clad >= -30, Months < 26) %>%
  mutate(Value = ifelse(grepl("No", Value), "Not taken", Value))

# how many switched between aza and mmf of the entire cohort
switch_patients <- meds %>% 
  filter(Immunosuppression %in% c("Mycophenolate-Mofetil", "Azathioprine"), Value != "Not taken") %>%
  group_by(Record.ID) %>%
  dplyr::count(Value) %>%
  dplyr::count(Record.ID) %>%
  filter(n >= 2) %>%
  pull(Record.ID)
  
switch_started_aza <- meds %>% 
  filter(Immunosuppression %in% c("Mycophenolate-Mofetil", "Azathioprine"), Value != "Not taken") %>%
  filter(Record.ID %in% switch_patients) %>%
  arrange(Record.ID, Days) %>%
  dplyr::select(Record.ID, Immunosuppression) %>%
  distinct() %>%
  group_by(Record.ID) %>%
  slice_head(., n = 1) %>%
  filter(Immunosuppression == "Azathioprine") %>%
  pull(Record.ID)
length(switch_started_aza) # 16

switch_started_mmf <- meds %>% 
  filter(Immunosuppression %in% c("Mycophenolate-Mofetil", "Azathioprine"), Value != "Not taken") %>%
  filter(Record.ID %in% switch_patients) %>%
  arrange(Record.ID, Days) %>%
  dplyr::select(Record.ID, Immunosuppression) %>%
  distinct() %>%
  group_by(Record.ID) %>%
  slice_head(., n = 1) %>%
  filter(Immunosuppression == "Mycophenolate-Mofetil") %>%
  pull(Record.ID)
length(switch_started_mmf) # 4

16/69*100
```

```{r}
a <- meds %>% 
  filter(Immunosuppression %in% c("Prednisolone"))

p <- 
  ggplot(a, aes(x = Months, y = Record.ID)) +
    geom_path(aes(group = Record.ID), colour = "grey", size = 1) +
    geom_point(aes(fill = Value), shape = 21) +
    geom_point(data = a %>% filter(Record.ID %in% biomarker_patients), aes(x = round(clad_day_clinical/30.417, 1), y = Record.ID), shape = 4, colour = "red") +
    facet_grid(Study+Group~Immunosuppression, scales = "free", space = "free", switch = "y") +
    theme +
    theme(axis.text.y = element_blank(), legend.position = "bottom") +
    labs(title = "Corticosteroids", y = "Patient", fill = "") +
  scale_fill_manual(values = c("Prednisolone" = "purple",  "Not taken" = "white"))
ggsave(paste0(analysis_path, "clinical-metadata/prednisolone.png"), p, width = 10, height = 14, units = 'cm', dpi = 600)

a <- meds %>% 
  filter(Immunosuppression %in% c("Tacrolimus", "Cyclosporine", "Everolimus"))

p <-
  ggplot(a, aes(x = Months, y = Record.ID)) +
    geom_path(aes(group = Record.ID), colour = "grey", size = 1) +
    geom_point(data = a %>% filter(Record.ID %in% biomarker_patients), aes(x = round(clad_day_clinical/30.417, 1), y = Record.ID), shape = 4, colour = "red", alpha = 0.6) +
    geom_point(aes(fill = Value), shape = 21) +
    facet_grid(Study+Group~Immunosuppression, scales = "free", space = "free", switch = "y") +
    theme +
    theme(axis.text.y = element_blank(), legend.position = "bottom") +
    labs(title = "Calcineurin and mTOR inhibitors", y = "Patient", fill = "") +
  scale_fill_manual(values = c("Tacrolimus" = "blue", "Cyclosporine" = "orange", "Not taken" = "white", "Everolimus" = "purple"))

ggsave(paste0(analysis_path, "clinical-metadata/tacro-cyclo-evero.png"), p, width = 17, height = 14, units = 'cm', dpi = 600)

a <- meds %>% 
  filter(Immunosuppression %in% c("Mycophenolate-Mofetil", "Azathioprine")) %>%
  mutate(Switched = case_when(Record.ID %in% switch_patients ~ "MMF or\nAza start",
                              TRUE ~ "No switch"))

p <- 
  ggplot(a, aes(x = Months, y = Record.ID)) +
    geom_path(aes(group = Record.ID), colour = "grey", size = 1) +
    geom_point(aes(fill = Value), shape = 21) +
    geom_point(data = a %>% filter(Record.ID %in% biomarker_patients), aes(x = round(clad_day_clinical/30.417, 1), y = Record.ID), shape = 4, colour = "red") +
    facet_grid(Study+Group+Switched~Immunosuppression, scales = "free", space = "free", switch = "y") +
    theme +
    theme(axis.text.y = element_blank(), legend.position = "bottom") +
    labs(title = "Antiproliferative medication", y = "Patient", fill = "") +
  scale_fill_manual(values = c("Mycophenolate-Mofetil" = "darkgreen", "Azathioprine" = "orange", "Not taken" = "white"))

  ggsave(paste0(analysis_path, "clinical-metadata/mmf-aza.png"), p, width = 16, height = 17, units = 'cm', dpi = 600)
```


# -------------------------------

## HLA mismatch

```{r}
hla <- read_csv("~/Desktop/R/Biomarker2/Redcap/Data/hla/hla-assessment.csv", col_types = "nccnnnccnnnnnnnnnnncccccc") %>% 
  filter(Record.ID %in% biomarker_patients) %>% 
  left_join(clad_assessment, by = "Record.ID")

hla.list <- list()

# Class I
hla.list[["All-Cl1"]] <-  hla %>% ggplot(aes(x = Group, y = `All Cl1 Eplets`)) +
  labs(title = "All Cl1 Eplets",
       y = "Number of mismatches") +
  theme +
  theme(axis.title.x = element_blank()) +
  stat_compare_means(label = "p.signif", method = "t.test", label.x.npc = "centre", label.y = 30 ) +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(aes(fill = Group), width = 0.1, shape = 21) +
  plot_fill(Group = TRUE) +
  #scale_y_continuous(limits = c(0,35)) +
  theme(legend.position = "bottom", legend.justification = c("left"))


hla.list[["All-Cl2"]] <- hla %>% ggplot(aes(x = Group, y = `Cl2 # eps all`)) +
    labs(title = "All Cl2 Eplets",
       y = "Number of mismatches") +
    theme +
    theme(axis.title.x = element_blank()) +
  stat_compare_means(label = "p.signif", method = "t.test",
                     label.x.npc = "centre",
                     label.y = 35) +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(aes(fill = Group), width = 0.1, shape = 21) +
  plot_fill(Group = TRUE) +
  #scale_y_continuous(limits = c(0,35)) +
  theme(legend.position = "bottom", legend.justification = c("left"))


# Class II
hla.list[["All-drb"]] <- hla %>% ggplot(aes(x = Group, y = `allDRB`)) +
    labs(title = "All DRB Eplets",
       y = "Number of mismatches") +
    theme +
    theme(axis.title.x = element_blank()) +
  stat_compare_means(label = "p.signif", method = "t.test",
                     label.y = 35,
                     label.x.npc = "centre") +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(aes(fill = Group), width = 0.1, shape = 21) +
  plot_fill(Group = TRUE) +
  #scale_y_continuous(limits = c(0,40)) +
  theme(legend.position = "bottom", legend.justification = c("left"))


hla.list[["All-dqb"]] <- hla %>% ggplot(aes(x = Group, y = `allDQB`)) +
    labs(title = "All DQB",
       y = "Number of mismatches") +
    theme +
    theme(axis.title.x = element_blank()) +
  stat_compare_means(label = "p.signif", method = "t.test",
                     label.y = 20, label.x.npc = "centre") +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(aes(fill = Group), width = 0.1, shape = 21) +
  plot_fill(Group = TRUE) +
  #scale_y_continuous(limits = c(0,25)) +
  theme(legend.position = "bottom", legend.justification = c("left"))


hla.list[["All-dqa"]] <- hla %>% ggplot(aes(x = Group, y = `allDQA`)) +
    labs(title = "All DQA",
       y = "Number of mismatches") +
    theme +
    theme(axis.title.x = element_blank()) +
  stat_compare_means(label = "p.signif", method = "t.test",
                     label.y = 6, label.x.npc = "centre") +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(aes(fill = Group), width = 0.1, shape = 21) +
  plot_fill(Group = TRUE) +
  #scale_y_continuous(limits = c(0,10)) +
  theme(legend.position = "bottom", legend.justification = c("left"))

p <- ggarrange(hla.list$`All-Cl1`, hla.list$`All-Cl2`, 
          hla.list$`All-drb`, hla.list$`All-dqb`, hla.list$`All-dqa`,
          nrow = 1,
          common.legend = TRUE, legend = "none")

ggsave("hla-mismatch.pdf", p, width = 26, height = 5, units = 'cm')
```


## Hospitalisations

```{r}
hospitalisations <- master_metadata$hosp_metadata %>%
  filter(Record.ID %in% biomarker_patients) %>% 
  inner_join(clad_assessment, by = "Record.ID") %>%
  distinct() %>%
  mutate(Admission_indication = gsub("GIT", "Gastrointestinal", Admission_indication),
         Group = factor(Group, levels = c("CLAD-free", "CLAD")),
         days_to_clad = clad_day_clinical - Days,
         Status = ifelse(days_to_clad > 0 | is.na(days_to_clad), "(before)", "(after)"),
         Admission_indication_status = paste(Admission_indication, Status, sep = " "),
         CLAD_diagnosis = "CLAD diagnosis",
         Months = round(Days/30.417, 1)) %>%
  filter(is.na(days_to_clad) | days_to_clad >= -30, Months < 26) 

# Days
p <-
 
  ggplot(hospitalisations, aes(x = round(Days/30.417, 1), y = as.character(Record.ID))) +
      geom_path(aes(group = Record.ID)) +
      geom_point(aes(fill = Admission_indication_status), shape = 21, size = 1.5) +
      
  geom_point(data = hospitalisations %>% filter(Group == "CLAD"), aes(x = round(clad_day_clinical/30.417, 1), y = as.character(Record.ID), colour = CLAD_diagnosis), shape = 4, size = 1) + #clad day
  
  labs(title = "Hospitalisations",
       y = "Patient",
       x = "Months",
       colour = "",
       fill = "Indication") +
  theme +
  theme(axis.text.y = element_blank()) +
  facet_grid(Group~., scales = "free_y", space = "free_y", switch = "y") +
  plot_fill(Admission_indication_status = TRUE) +
  scale_colour_manual(values = c("CLAD diagnosis" = "red")) 

ggsave("hospitalisations.pdf", p, width = 12, height = 8, units = 'cm')


# Alluvium
d <- hospitalisations %>%
  mutate(Status = ifelse(days_to_clad > 0 | is.na(days_to_clad), "(before)", "(after)"),
         Admission_indication_status = paste(Admission_indication, Status, sep = " "),
         Months_grouped = case_when(Days <= 42 ~ "<1.5",
                                    Days > 42 & Days <= 92 ~ "1.5-3",
                                    Days > 92 & Days <= 183 ~ "4-6",
                                    Days > 183 & Days <= 273 ~ "7-9",
                                    Days > 273 & Days <= 365 ~ "10-12",
                                    Days > 365 & Days <= 547 ~ "13-18",
                                    Days > 547 ~ ">18"), 
         Months_grouped = factor(Months_grouped, c("<1.5", "1.5-3", "4-6", "7-9", "10-12", "13-18", ">18")) ) %>%
  group_by(Months_grouped, Group) %>% 
  dplyr::count(Admission_indication_status) %>%
  mutate(Percentage = round(n/sum(n)*100, 2)) 
  
p <- 
  ggplot(d, aes(x = Months_grouped, y = Percentage, fill = Admission_indication_status, 
                alluvium = Admission_indication_status, stratum = Admission_indication_status)) +
    geom_alluvium(width = 0.6) +
    geom_stratum(width = 0.6) +
    labs(title = "Hospitalisations", x = "Time post-transplant", y = "Percentage (%)", fill = "Indication") +
    theme +
    facet_grid(Group~.) + #switch = "y"
    plot_fill(Admission_indication_status = TRUE) 

ggsave("hospitalisations-alluvium.pdf", p, width = 16, height = 8, units = 'cm')
```


## Neutrophils in blood

```{r}
blood_cells <- master_metadata$bronch_metadata %>%
  dplyr::select(Record.ID, Days, Neutrophils, WCC, Lymphocytes) %>%
  filter(Record.ID %in% biomarker_patients) %>% 
  inner_join(clad_assessment, by = "Record.ID") %>%
  distinct() %>%
  mutate(Group = factor(Group, levels = c("CLAD-free", "CLAD")),
         days_to_clad = clad_day_clinical - Days,
         Status = ifelse(days_to_clad > 0 | is.na(days_to_clad), "(before)", "(after)"),
         CLAD_diagnosis = "CLAD diagnosis",
         Months = round(Days/30.417, 1)) %>%
  filter(is.na(days_to_clad) | days_to_clad >= -30, Months < 26) 

p <- 
  
  ggplot(blood_cells, aes(x = Months, y = WCC)) +
  theme +
  labs(x = "Months post-transplant", 
       y = "Neutrophils", 
       title = "Blood immune cells") +
  geom_smooth(aes(group = Record.ID), method = "lm", formula = y ~ ns(x, 3), se = FALSE, colour = "lightgrey", linewidth = 0.3) +
  geom_smooth(aes(group = Group, colour = Group), method = "lm", formula = y ~ ns(x, 3), linewidth = 1 ) +
  scale_x_continuous(breaks = c(3, 6, 12, 18, 24)) +
  geom_vline(xintercept = 12, lty = "dashed") +
  plot_colours(Group = TRUE) +
  plot_fill(Group = TRUE)

ggsave("blood-cells.pdf", p, width = 10, height = 7, units = 'cm')



```



# Antibiotics


```{r}
# Alluvium
antibiotics <- master_metadata$meds_metadata %>%
  dplyr::select(Patient_days, blactamother, `sulfonamide-benzylpyrimidine`, macrolide) %>%
  separate(Patient_days, c("Record.ID", "Days"), sep = "_") %>%
  mutate(Days = as.numeric(Days) , # ignore NA message
         Record.ID = as.numeric(Record.ID), 
         blactamother = ifelse(blactamother > 0, "Yes", "No"),
         `sulfonamide-benzylpyrimidine` = ifelse(`sulfonamide-benzylpyrimidine` > 0, "Yes", "No"),
         macrolide = ifelse(macrolide > 0, "Yes", "No") ) %>%
  filter(Record.ID %in% biomarker_patients) %>%
  pivot_longer(cols = blactamother:macrolide, names_to = "Antibiotic", values_to = "Taken") %>%
  mutate(Antibiotic = gsub("blactamother", "B-lactam & others", Antibiotic),
         Antibiotic = gsub("macrolide", "Macrolide", Antibiotic),
         Antibiotic = gsub("sulfonamide-benzylpyrimidine", "Sulfonamide-benzylpyrimidine", Antibiotic) ) %>%
  left_join(clad_assessment) %>%
  mutate(days_to_clad = clad_day_clinical - Days,
         Status = ifelse(days_to_clad > 0 | is.na(days_to_clad), "(before)", "(after)"),
         Antibiotic_status = paste(Antibiotic, Status, sep = " "),
         Time_interval = case_when(Days >= 0 & Days < 42 ~ "0-6W",
                                   Days >= 42 & Days < 92 ~ "7W-3M",
                                   Days >= 92 & Days < 183 ~ "4-6M",
                                   Days >= 183 & Days < 365 ~ "7-12M",
                                   Days >= 365 & Days < 548 ~ "13-18M",
                                   Days >= 548 ~ "19-30M" ),
         Time_interval = factor(Time_interval, levels = c("0-6W", "7W-3M", "4-6M", "7-12M", "13-18M", "19-30M"))) 


d <- antibiotics %>%
  filter(Taken == "Yes") %>%
  group_by(Time_interval, Group) %>% 
  dplyr::count(Antibiotic) %>%
  mutate(Percentage = round(n/sum(n)*100, 2)) 

p <- 
  ggplot(d, aes(x = Time_interval, y = Percentage, fill = Antibiotic, 
                alluvium = Antibiotic, stratum = Antibiotic)) +
    geom_alluvium(width = 0.6) +
    geom_stratum(width = 0.6) +
    labs(title = "Antibiotic", x = "Time post-transplant", y = "Percentage (%)", fill = "Indication") +
    theme +
    facet_grid(Group~.) + #switch = "y"
    plot_fill(Admission_indication_status = TRUE) 

ggsave("hospitalisations-alluvium.pdf", p, width = 16, height = 8, units = 'cm')



# Pie
d <- antibiotics %>%
  group_by(Time_interval, Group, Antibiotic) %>% 
  dplyr::count(Taken) %>%
  mutate(Percentage = round(n/sum(n)*100, 2),
         Taken = factor(Taken),
         ymax = cumsum(Percentage),
         ymin = c(0, head(ymax, n = -1)) )
  
p <- 
  
  ggplot(d %>% filter(Antibiotic == "B-lactam & others"), aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill = Taken)) +
     geom_rect(alpha = 0.8) +
     coord_polar(theta = "y") +
  xlim(c(2, 4)) +
  labs(title = "Antibiotics post-transplant",
       x = "Time post-transplant",
       y = "Percentage (%)") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), strip.text = element_text(face = "bold"),
        legend.position = "bottom", legend.justification = "centre") +
  labs(fill = "Antibiotics") +
  facet_wrap(Group~Time_interval, ncol = 6) #+
  #scale_fill_manual(values = c("lightgrey", "black", "lightblue", "blue"))

ggsave("antibiotics.pdf", p, width = 10, height = 8, units = 'cm')
```


```{r}
d <- data.frame(sample_data(bact.logcss)) %>%
  group_by(Days_interval) %>%
  count(sulfbenz_cat) %>%
  mutate(Percentage = n/sum(n)*100,
         sulfbenz_cat = factor(sulfbenz_cat),
         ymax = cumsum(Percentage),
         ymin = c(0, head(ymax, n = -1)))
  
p <- 
  
  ggplot(d, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill = sulfbenz_cat)) +
     geom_rect(alpha = 0.8) +
     coord_polar(theta = "y") +
  xlim(c(2, 4)) +
  labs(title = "Sulfonamide-benzylpyrimidine post-transplant",
       x = "Days interval",
       y = "Percentage") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), strip.text = element_text(face = "bold"),
        legend.position = "bottom", legend.justification = "centre") +
  labs(fill = "BactrimDS") +
  facet_wrap(~Days_interval) +
  scale_fill_manual(values = c("lightgrey", "black", "lightblue", "blue"))

ggsave("clinical-checks/sulfonamide-benzylpyrimidine.png", p, width = 10, height = 8, units = 'cm')


d <- data.frame(sample_data(bact.logcss)) %>%
  group_by(Days_interval) %>%
  count(macrolide_cat) %>%
  mutate(Percentage = n/sum(n)*100,
         macrolide_cat = factor(macrolide_cat),
         ymax = cumsum(Percentage),
         ymin = c(0, head(ymax, n = -1)))
  
p <- 
  
  ggplot(d, aes(ymax=ymax, ymin=ymin, xmax=4, xmin=3, fill = macrolide_cat)) +
     geom_rect(alpha = 0.8) +
     coord_polar(theta = "y") +
  xlim(c(2, 4)) +
  labs(title = "Macrolide post-transplant",
       x = "Days interval",
       y = "Percentage") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5), strip.text = element_text(face = "bold"),
        legend.position = "bottom", legend.justification = "centre") +
  labs(fill = "Azythromycin") +
  facet_wrap(~Days_interval) +
  scale_fill_manual(values = c("lightgrey", "black", "lightblue", "blue"))

ggsave("clinical-checks/macrolide.png", p, width = 10, height = 8, units = 'cm')
```

# Medications average

```{r}
# Medications on sversge considering redcap metadata, not dataset available
meds <- master.metadata$bronch_metadata %>%
  filter(Record.ID %in% unique(datasets[[1]]$Record.ID)) %>% 
  dplyr::select(Record.ID, Days, contains(c("Posaconazole", "Voriconazole", "AmphothericinB", "Ciprofloxacin", "Piperacillin-Tazobactam", "Augmentin", "Meropenem", "Tobramycin", "Pentamidine", "Moxifloxacin", "Colistin", "Resprim", "Vancomycin", "Amoxycillin", "Isozianid", "Ceftazidime", "Flucloxacillin", "Tacrolimus", "Cyclosporin", "Everolimus", "Azathioprine", "Mycophenolate-Mofetil", "Prednisolone", "BactrimDS", "Valganciclovir", "Ganciclovir", "Azithromycin",  "Valaciclovir"))) %>%
  tidyr::unite(col = "Medication", colnames(.)[3:31], sep = ",", na.rm = TRUE) %>%
  separate_rows(Medication, sep = ",") %>%
  mutate(Taken = ifelse(str_detect(Medication, "No"), "No", "Yes"),
         Medication = gsub("No", "", Medication)) %>%
  right_join(meds_list %>% rename(Medication = Name, Group_action2 = Group), by = "Medication") 
```

```{r}
# Other and b-lactams
p1 <- 
  
  meds %>% 
  filter(Group_action %in% c("b-lactam", "other")) %>%
  filter(Taken == "Yes") %>%
  full_join(distinct(meds[, c("Record.ID", "Days")])) %>%
  
  ggplot(aes(x = Months, y = as.character(Record.ID))) +
  geom_tile(aes(fill = Group_action, colour = BAL_class_combined), size=0.3) +
  labs(title = "B-lactams and miscellaneous antibiotics",
       fill = "Type",
       colour = "BAL class",
       x = "Months post-transplant",
       y = "Patient") +
  theme +
  theme(axis.text.y = element_blank()) +
  guides(colour = "none") +
  scale_x_continuous(expand=c(0, 0), breaks = c(3, 6, 12, 24, 36)) +
  scale_fill_manual(values = c("b-lactam" = "darkgrey", "other" = "#FA8128")) 

## Bactrim DS
p2 <-
  meds %>% 
  filter(Group_action %in% c("sulfonamide-benzylpyrimidine")) %>%
  filter(Taken == "Yes") %>%
  full_join(distinct(meds[, c("Record.ID", "Days")])) %>%
  
  ggplot(aes(x = Months, y = as.character(Record.ID))) +
  geom_tile(aes(fill = Medication, colour = BAL_class_combined), fill = "darkgrey", size=0.3) +
  labs(title = "BactrimDS",
       colour = "BAL class",
       fill = "Medication",
       x = "Months post-transplant",
       y = "Patient") +
  theme +
  theme(axis.text.y = element_blank()) +
  #guides(colour = "none") +
  scale_x_continuous(expand=c(0, 0), breaks = c(3, 6, 12, 24, 36)) 
  
## Macrolides
p3 <- 
  meds %>% 
  filter(Group_action %in% c("macrolide")) %>%
  filter(Taken == "Yes") %>%
  full_join(distinct(meds[, c("Record.ID", "Days")])) %>%

  ggplot(aes(x = Months, y = as.character(Record.ID))) +
  geom_tile(aes(colour = BAL_class_combined), fill = "darkgrey", size=0.3) +
  labs(title = "Azithromycin",
       fill = "Type",
       colour = "BAL class",
       x = "Months post-transplant",
       y = "Patient") +
  theme +
  theme(axis.text.y = element_blank()) +
  scale_x_continuous(expand=c(0, 0), breaks = c(3, 6, 12, 24, 36)) 

ggsave("antibiotics.pdf", p, width = 12, height = 24, units = 'cm')
```

```{r}
# Antifungals
p4 <- 
  
  meds %>% 
  filter(Type %in% c("Antifungal")) %>%
  filter(Taken == "Yes") %>%
  full_join(distinct(meds[, c("Record.ID", "Days")])) %>%

  ggplot(aes(x = Months, y = as.character(Record.ID))) +
  geom_tile(aes(fill = Medication, colour = BAL_class_combined), size=0.3) +
  labs(title = "Posaconazole, Voriconazole and AmphothericinB",
       x = "Months post-transplant",
       y = "Patient") +
  theme +
  theme(axis.text.y = element_blank()) +
  guides(colour = "none") +
  scale_x_continuous(expand=c(0, 0), breaks = c(3, 6, 12, 24, 36)) +
  scale_fill_manual(values = c("Posaconazole" = "grey", "AmphothericinB" = "#FA8128", "Voriconazole" = "#FCAE1E")) 


# Antivirals
p5 <- 
  
  meds %>% 
  filter(Type %in% c("Antiviral")) %>%
  dplyr::select(!Action) %>%
  filter(Taken == "Yes") %>%
  full_join(distinct(meds[, c("Record.ID", "Days")])) %>%
  
  ggplot(aes(x = Months, y = as.character(Record.ID))) +
  geom_tile(aes(fill = Medication, colour = BAL_class_combined), size=0.3) +
  labs(title = "Valganciclovir, Ganciclovir and Valaciclovir",
       x = "Months post-transplant",
       y = "Patient") +
  theme +
  theme(axis.text.y = element_blank()) +
  guides(colour = "none") +
  scale_x_continuous(expand=c(0, 0), breaks = c(3, 6, 12, 24, 36)) +
  scale_fill_manual(values = c("Valganciclovir" = "grey", "Valaciclovir" = "#FA8128", "Ganciclovir" = "#FCAE1E")) 
  
p <- ggarrange(p4, p5, ncol = 1, legend = "right", align = "v")

ggsave("clinical-checks/antifungals-antivirals.pdf", p, width = 12, height = 16, units = 'cm')
```


```{r}
p6 <- 
    meds %>% 
  filter(Type %in% c("Steroid")) %>%
  filter(Taken == "Yes") %>%
  full_join(distinct(meds[, c("Record.ID", "Days")])) %>%
  
  ggplot(aes(x = Months, y = as.character(Record.ID))) +
  geom_tile(aes(colour = BAL_class_combined), fill = "darkgrey", size=0.3) +
  labs(title = "Prednisolone",
       fill = "Type",
       colour = "BAL class",
       x = "Months post-transplant",
       y = "Patient") +
  theme +
  theme(axis.text.y = element_blank()) +
  scale_x_continuous(expand=c(0, 0), breaks = c(3, 6, 12, 24, 36)) +


p7 <- 
  meds %>% 
  filter(Medication %in% c("Azathioprine", "Mycophenolate-Mofetil")) %>%
  filter(Taken == "Yes") %>%
  full_join(distinct(meds[, c("Record.ID", "Days")])) %>%
  
  ggplot(aes(x = Months, y = as.character(Record.ID))) +
  geom_tile(aes(fill = Medication, colour = BAL_class_combined), size=0.3) +
  labs(title = "Azathioprine and Mycophenolate-Mofetil",
       fill = "Type",
       colour = "BAL class",
       x = "Months post-transplant",
       y = "Patient") +
  theme +
  theme(axis.text.y = element_blank()) +
  guides(colour = "none") +
  scale_x_continuous(expand=c(0, 0), breaks = c(3, 6, 12, 24, 36)) +
  scale_fill_manual(values = c("Azathioprine" = "grey", "Mycophenolate-Mofetil" = "#FA8128")) 


p8 <- 
  meds %>% 
  filter(Medication %in% c("Tacrolimus", "Everolimus", "Cyclosporin")) %>%
  filter(Taken == "Yes") %>%
  full_join(distinct(meds[, c("Record.ID", "Days")])) %>%
  
  ggplot(aes(x = Months, y = as.character(Record.ID))) +
  geom_tile(aes(fill = Medication, colour = BAL_class_combined), size=0.3) +
  labs(title = "Tacrolimus, Cyclosporin and Everolimus",
       fill = "Type",
       colour = "BAL class",
       x = "Months post-transplant",
       y = "Patient") +
  theme +
  theme(axis.text.y = element_blank()) +
  guides(colour = "none") +
  scale_x_continuous(expand=c(0, 0), breaks = c(3, 6, 12, 24, 36)) +
  scale_fill_manual(values = c("Tacrolimus" = "grey", "Cyclosporin" = "#FA8128", "Everolimus" = "#FCAE1E")) 


p <- ggarrange(p6, p7, p8, ncol = 1, legend = "right", align = "v")

ggsave("immunosuppressants.pdf", p, width = 12, height = 24, units = 'cm')
```

```{r}
##How many started on MMF then switched to Azathioprine

  meds %>% 
  filter(Medication %in% c("Azathioprine", "Mycophenolate-Mofetil"), Taken == "Yes") %>%
  dplyr::select(Record.ID, Days, BAL_class_combined, Medication) %>%
  group_by(Record.ID, Medication) %>%
  slice_min(Days, n = 2) %>%
  group_by(Record.ID) %>%
  arrange(Record.ID, Days) %>%
  mutate(Ever_changed = case_when(length(unique(Medication)) == 1 ~ "Never",
                                 length(unique(Medication)) > 1 ~ "Ever"),
         Switched_from = ifelse(Ever_changed == "Ever", unique(Medication)[1], "None"),
         ) %>%
  group_by(Ever_changed, Switched_from) %>%
  count(Record.ID) %>%
  group_by(Ever_changed, Switched_from) %>%
  dplyr::count() %>%
  ungroup() %>%
  #group_by(Ever_changed) %>%
  mutate(Percentage = round(n/sum(n)*100, 2)) 
  
  
  ggplot(aes(x = Days, y = as.factor(Record.ID))) +
  geom_path(aes(group = Record.ID), colour = "grey", size = 0.5) +
  geom_point(aes(fill = Medication, shape = BAL_class_combined)) +
  theme +
  scale_shape_manual(values = c(21, 24)) +
  facet_grid(Ever_changed~., scales = "free_y", space = "free_y")
```

```{r}
# Antibiotics
#p <- 
  
  meds %>% 
  filter(Group_action %in% c("sulfonamide-benzylpyrimidine")) %>%
  dplyr::select(!Action) %>%
  pivot_wider(names_from = Medication, values_from = Taken) %>%
  mutate(Taken = ifelse(BactrimDS == "Yes" | Trimethoprim == "Yes", "Yes", "No")) %>%
  group_by(Time_interval_detailed) %>% 
  count(Taken) %>%
  mutate(n = round(n/sum(n)*100, 2)) %>%
  rename(Count = n) %>%
  
  ggplot(aes(x = Time_interval_detailed, y = Count, fill = Taken)) +
  geom_bar(stat = 'identity', position = 'fill') +
  labs(title = "BactrimDS and trimethoprim antibiotics at bronchoscopy",
       x = "Months post-transplant",
       y = "Patient") +
  theme

#ggsave(paste0("clinical-checks/antibiotics.png"), p, width = 12, height = 14, units = 'cm')
```

```{r}
# Average medications over time
meds.time <- meds %>%
  filter(Type == "Antibiotic") %>%
  group_by(Months, Group, Record.ID) %>%
  dplyr::count() %>% 
  rename(Count = n) %>%
  #left_join(meds_list %>% rename(Medication = Name, Group_med = Group), by = "Medication") %>%
  mutate(Count_char = case_when(Count < 2 ~ "1",
                           Count >= 2 & Count < 5 ~ "2-4",
                           Count >= 5 & Count < 10 ~ "5-9",
                           Count >= 10 & Count < 20 ~ "10-19",
                           Count >= 20 ~ ">20"),
         Count_char = factor(Count_char, levels = c("1", "2-4", "5-9", "10-19",">20"))) 

p <- 
  
  ggplot(meds.time, aes(x = Months, y = Group, fill = Count_char)) +
  geom_tile(colour="white", size=0.5) +
  labs(title = "Antibiotics at brochoscopy",
       x = "Months post-transplant",
       fill = "Count per patient") +
  theme + 
  #facet_grid(Type~., scales = "free_y", space = "free_y") +
  scale_y_discrete(expand=c(0, 0)) +
  #scale_x_discrete(expand=c(0, 0)) +
  scale_fill_viridis(direction = -1) +
    theme(legend.position = "bottom", legend.justification = c("left"))

ggsave(paste0("clinical-checks/medications-time.png"), p, width = 22, height = 14, units = 'cm')
## rename columns
  #setNames(c("year", "week", "state", "value"))


meds %>%
  filter(Type == "Antibiotic") %>%
  group_by(Interval_dosage2, Group_med, Record.ID) %>%
  dplyr::count() %>%
  group_by(Interval_dosage2) %>%
  count(Group_med) %>%
  mutate(Percentage = n/sum(n)*100)

  ggplot(meds, aes(x = Months, y = Group, fill = Count_char)) +
  geom_tile(colour="white", size=0.5) +
  labs(title = "Antibiotics at brochoscopy",
       x = "Months post-transplant",
       fill = "Count per patient") +
  theme + 
  #facet_grid(Type~., scales = "free_y", space = "free_y") +
  scale_y_discrete(expand=c(0, 0)) +
  #scale_x_discrete(expand=c(0, 0)) +
  scale_fill_viridis(direction = -1) +
    theme(legend.position = "bottom", legend.justification = c("left"))
```


### BAL microbiology
#### Bacterial

```{r}
micro.bact <- master.metadata$bronch_metadata %>%
    filter(Record.ID %in% patients) %>%
  dplyr::select(Record.ID, Days, BronchoalveolarLavageCulture, NotGrowth, StaphylococcusAureusMRSA, PseudomonasSpecies, KlebsiellaPneumoniae, SerratiaMarcescens, StenotrophomonasMaltophilia, EscherichiaColi, EnterobacterAerogenes) %>%
  filter(BronchoalveolarLavageCulture != "Not Done") %>%
  left_join(clad.blad.assessment, by = "Record.ID") 

micro.bact.cul <- micro.bact %>% 
  group_by(Group, Record.ID, BronchoalveolarLavageCulture) %>%
  dplyr::count() %>% 
  rename(Count = n) %>%
  mutate(Culture = case_when(BronchoalveolarLavageCulture == "0" ~ "Negative",
                             BronchoalveolarLavageCulture == "1" ~ "Positive"),
         Group = factor(Group, levels = c("CLAD-free", "CLAD")))

# Percentage of positive cultures per group
l <- list()
for (i in unique(micro.bact.cul$Record.ID)){
  
  f <- micro.bact.cul %>% 
    filter(Record.ID == i) %>%
    arrange(Culture)
  
  # If only one row
  if (dim(f)[1] < 2){
    # If Positive only
    if (f$Culture %in% "Positive"){
      f <- f %>%
        mutate(Percentage = round(100*(.$Count[1]/.$Count[1])))}
    # If Negative only
    if (f$Culture %in% "Negative"){
      f <- f %>%
        mutate(Percentage = round(100*(0/.$Count[1])))}}
  
  if (dim(f)[1] == 2){
    f <- f %>%
      mutate(Percentage = round(100*(.$Count[2]/(.$Count[1] + .$Count[2]))))} #Neg/Neg+Pos
  
  l[[i]] <- f}

micro.bact.cul <- as.data.frame(bind_rows(l))

p <- 
  
  micro.bact.cul %>%
  filter(Culture == "Positive") %>%
  
  ggplot(aes(x = Group, y = Percentage)) +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(shape = 21, width = 0.2) +
  labs(title = "Positive BAL bacterial cultures",
       y = "Percentage per patient") +
  theme +
  theme(axis.title.x = element_blank()) +
  theme(legend.position = "bottom", legend.justification = c("left")) +
  stat_compare_means(label = "p.signif", 
                     method = "wilcox.test",
                     label.x.npc = "centre") +
  plot_fill(Group = TRUE)
ggsave("clinical-checks/micro-bact-cul.pdf", p, width = 9, height = 9, units = 'cm') # ns
  
 
# Positive micro species
micro.bact.pos <- micro.bact %>%
  filter(BronchoalveolarLavageCulture == "1") %>%
  dplyr::select(!NotGrowth) %>%
  unite(col = "Species", colnames(.)[4:10], sep = ",", na.rm = TRUE) %>%
  separate_rows(Species, sep = ",") %>%
  filter(!grepl("No", Species)) %>%
  mutate(Patient_days = paste0(Record.ID, "_", Days)) %>%
  group_by(Group, Patient_days, Species) %>%
  dplyr::count() %>% 
  rename(Count = n) %>%
  mutate(Group = factor(Group, levels = c("CLAD-free", "CLAD")))

name <- "micro-bact-pos"
p <- 
  
  micro.bact.pos %>%
  ggplot(aes(x = Patient_days, y = Count, fill = Species)) +
  geom_bar(stat = 'identity', position = 'fill') +
  labs(title = "Positive BAL bacterial cultures - composition",
       x = "Sample",
       y = "Percentage") +
  theme +
  theme(axis.text.x = element_blank(),
        strip.background =element_rect(fill = "white"),
        strip.text = element_text(face = "bold", colour = "black")) +
  facet_grid(.~Group, scales = "free_x", space = "free_x")
ggsave(paste0("clinical-checks/", name, ".pdf"), p, width = 16, height = 8, units = 'cm') # ns

name <- "micro-bact-pos-count"
p <- 
  
micro.bact.pos %>%
  
  ggplot(aes(x = Group, y = Count)) +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(shape = 21, width = 0.2) +
  labs(title = "Bacterial cultures",
       y = "Count per patient") +
  theme +
  theme(axis.title.x = element_blank(),
        strip.background =element_rect(fill = "white"),
        strip.text = element_text(face = "bold", colour = "black")) +
  theme(legend.position = "bottom", legend.justification = c("left")) +
  stat_compare_means(label = "p.signif", 
                     method = "wilcox.test",
                     #label.y = 4.5,
                     label.x.npc = "centre") +
  plot_fill(Group = TRUE) +
  facet_wrap(~Species, nrow = 1)
ggsave(paste0("clinical-checks/", name, ".pdf"), p, width = 24, height = 9, units = 'cm') # ns
```

#### Fungal

```{r}
micro.fung <- master.metadata$bronch_metadata %>%
    filter(Record.ID %in% patients) %>%
  dplyr::select(Record.ID, Days, BALFungalCulture, YeastSpecies, AspergillusSpecies, PenicilliumSpecies, ScedosporiumSpecies) %>%
  filter(BALFungalCulture != "Not Done") %>%
  left_join(clad.blad.assessment, by = "Record.ID") 

micro.fung.cul <- micro.fung %>% 
  group_by(Group, Record.ID, BALFungalCulture) %>%
  dplyr::count() %>% 
  rename(Count = n) %>%
  mutate(Culture = case_when(BALFungalCulture == "0" ~ "Negative",
                             BALFungalCulture %in% c("1 - Other") ~ "Positive",
                             BALFungalCulture %in% c("1 - Yeast Only") ~ "Positive"),
         Group = factor(Group, levels = c("CLAD-free", "CLAD")))

# Percentage of positive cultures per group
l <- list()
for (i in unique(micro.fung.cul$Record.ID)){
  
  f <- micro.fung.cul %>% 
    filter(Record.ID == i) %>%
    arrange(Culture)
  
  # If only one row
  if (dim(f)[1] < 2){
    # If Positive only
    if (f$Culture %in% "Positive"){
      f <- f %>%
        mutate(Percentage = round(100*(.$Count[1]/.$Count[1])))}
    # If Negative only
    if (f$Culture %in% "Negative"){
      f <- f %>%
        mutate(Percentage = round(100*(0/.$Count[1])))}}
  
  if (dim(f)[1] == 2){
    f <- f %>%
      mutate(Percentage = round(100*(.$Count[2]/(.$Count[1] + .$Count[2]))))} #Neg/Neg+Pos
  
  l[[i]] <- f}

micro.fung.cul <- as.data.frame(bind_rows(l))

name <- "micro-fung-cul"
p <- 
  
  micro.fung.cul %>%
  filter(Culture == "Positive") %>%
  
  ggplot(aes(x = Group, y = Percentage)) +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(shape = 21, width = 0.1) +
  labs(title = "Positive BAL fungal cultures",
       y = "Percentage per patient") +
  theme +
  theme(axis.title.x = element_blank()) +
  theme(legend.position = "bottom", legend.justification = c("left")) +
  stat_compare_means(label = "p.signif", 
                     method = "wilcox.test",
                     label.y = 99,
                     label.x.npc = "centre") +
  plot_fill(Group = TRUE)
ggsave(paste0("clinical-checks/", name, ".pdf"), p, width = 9, height = 9, units = 'cm') # 
  
 
# Positive micro species
micro.fung.pos <- micro.fung %>%
  filter(BALFungalCulture %in% c("1 - Other", "1 - Yeast Only")) %>%
  unite(col = "Species", colnames(.)[4:7], sep = ",", na.rm = TRUE) %>%
  separate_rows(Species, sep = ",") %>%
  filter(!grepl("No", Species)) %>%
  mutate(Patient_days = paste0(Record.ID, "_", Days)) %>%
  group_by(Group, Patient_days, Species) %>%
  dplyr::count() %>% 
  rename(Count = n) %>%
  mutate(Group = factor(Group, levels = c("CLAD-free", "CLAD")))

name <- "micro-fung-pos"
p <- 
  
  micro.fung.pos %>%
  ggplot(aes(x = Patient_days, y = Count, fill = Species)) +
  geom_bar(stat = 'identity', position = 'fill') +
  labs(title = "Positive BAL fungal cultures - composition",
       x = "Sample",
       y = "Count") +
  theme +
  theme(axis.text.x = element_blank(),
        strip.background =element_rect(fill = "white"),
        strip.text = element_text(face = "bold", colour = "black")) +
  facet_grid(.~Group, scales = "free_x", space = "free_x")
ggsave(paste0("clinical-checks/", name, ".pdf"), p, width = 16, height = 8, units = 'cm') # ns

name <- "micro-fung-pos-count"
p <- 
  
micro.fung.pos %>%
  
  ggplot(aes(x = Group, y = Count)) +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(shape = 21, width = 0.1) +
  labs(title = "Fungal cultures",
       y = "Count per patient") +
  theme +
  theme(axis.title.x = element_blank(),
        strip.background =element_rect(fill = "white"),
        strip.text = element_text(face = "bold", colour = "black")) +
  theme(legend.position = "bottom", legend.justification = c("left")) +
  stat_compare_means(label = "p.signif", 
                     method = "wilcox.test",
                     label.x.npc = "centre") +
  plot_fill(Group = TRUE) +
  facet_wrap(~Species)
ggsave(paste0("clinical-checks/",  name, ".pdf"), p, width = 12, height = 9, units = 'cm') # ns
```




```{r}
# custom sum function returns NA when all values in set are NA,
# in a set mixed with NAs, NAs are removed and remaining summed.
na_sum <- function(x){
  if(all(is.na(x))) val <- sum(x, na.rm=F)
  if(!all(is.na(x))) val <- sum(x, na.rm=T)
  return(val)}
```


### CMV mismatch and reactivation

```{r}
cmv.mismatch <- master.metadata$general_metadata %>%
  filter(Record.ID %in% patients) %>%
  dplyr::select(Record.ID, CMVMismatch, DonorCMV) %>%
  left_join(clad.blad.assessment, by = "Record.ID") %>%
  group_by(Group, CMVMismatch) %>%
  dplyr::count() %>%
  rename(Count = n)

# Percentage of positive cultures per group
l <- list()
for (i in unique(cmv.mismatch$Group)){
  f <- cmv.mismatch %>% 
    filter(Group == i) %>%
    arrange(desc(CMVMismatch)) %>%
      mutate(Percentage = round(100*(.$Count[2]/(.$Count[1] + .$Count[2])))) #Neg/Neg+Pos
  l[[i]] <- f}

cmv.mismatch <- as.data.frame(bind_rows(l))
```


```{r}
cmv <- master.metadata$bronch_metadata %>%
  filter(Record.ID %in% patients) %>%
  dplyr::select(Record.ID, CMVinBAL, IfPositiveCMVBALViralLoad, Days) %>%
  filter(CMVinBAL == "CMVinBAL") %>%
  left_join(clad.blad.assessment, by = "Record.ID") 
  
cmv.det <- cmv %>%
  group_by(Group, Record.ID, CMVinBAL) %>%
  dplyr::count()%>%
  rename(`Count` = n) %>%
  mutate(Group = factor(Group, levels = c("CLAD-free", "CLAD"))) 
  
# Percentage of positive cmv
l <- list()
for (i in unique(cmv.det$Record.ID)){
  
  f <- cmv.det %>% 
    filter(Record.ID == i) %>%
    arrange(desc(Result))
  
  # If only one row
  if (dim(f)[1] < 2){
    # If Positive only
    if (f$Result %in% "Detected"){
      f <- f %>%
        mutate(Percentage_positive = round(100*(.$Count[1]/.$Count[1])))}
    # If Negative only
    if (f$Result %in% "Not Detected"){
      f <- f %>%
        mutate(Percentage_positive = round(100*(0/.$Count[1])))}}
  
  if (dim(f)[1] == 2){
    f <- f %>%
      mutate(Percentage_positive = round(100*(.$Count[2]/(.$Count[1] + .$Count[2]))))} 
  
  l[[i]] <- f}

cmv.det <- as.data.frame(bind_rows(l))

# Positive detected
name <- "cmv-detected"
p <- 
  
  cmv.det %>%
  filter(Result == "Detected") %>%
  
  ggplot(aes(x = Group, y = Percentage_positive)) +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(shape = 21, width = 0.2) +
  labs(title = "CMV detected",
       y = "Percentage positive per patient") +
  theme +
  theme(axis.title.x = element_blank()) +
  theme(legend.position = "bottom", legend.justification = c("left")) +
  stat_compare_means(label = "p.signif", 
                     method = "wilcox.test",
                     label.x.npc = "centre") +
  plot_fill(Group = TRUE)
ggsave(paste0("clinical-checks/", name, ".pdf"), p, width = 9, height = 7, units = 'cm') # ns
```

### DSA mismatch

```{r}
dsa.mismatch <- master.metadata$general_metadata %>%
  filter(Record.ID %in% patients) %>%
  dplyr::select(Record.ID, DSAatTransplant) %>%
  left_join(clad.blad.assessment, by = "Record.ID") %>%
  group_by(Group, DSAatTransplant) %>%
  dplyr::count() %>%
  rename(Count = n)

# Percentage of positive per group
l <- list()
for (i in unique(dsa.mismatch$Group)){
  f <- dsa.mismatch %>% 
    filter(Group == i) %>%
    arrange(desc(DSAatTransplant)) %>%
      mutate(Percentage = round(100*(.$Count[2]/(.$Count[1] + .$Count[2])))) #Neg/Neg+Pos
  l[[i]] <- f}
dsa.mismatch <- as.data.frame(bind_rows(l))

# Detected over time
dsa.time <- clinical.metadata %>%
  filter(Repeat.Instrument == 'DSA Results', Record.ID %in% patients) %>%
  dplyr::select(Record.ID, Date.Of.DSAs:If.detected..DSA.details..including.MFI.) %>%
  filter(Date.Of.DSAs != "") %>%
  dplyr::select(Record.ID, "Days" = Time.From.Transplant.5, "Result" = DSA.Result) %>%
  left_join(clad.blad.assessment, by = "Record.ID") 

dsa.time %>%
  ggplot(aes(x = Days, y = as.character(Record.ID))) +
  geom_path(aes(group = Record.ID)) +
  geom_point(aes(colour = Result)) +
  theme +
  theme(axis.text.y = element_blank()) +
  theme(legend.position = "bottom", legend.justification = c("left")) +
  labs(title = "DSA over time",
       y = "Patient") +
  facet_grid(Group~., scales = "free_y", space = "free_y")
```


```{r}
## Positive detected
dsa.det <- dsa.time %>%
  group_by(Group, Record.ID, Result) %>%
  dplyr::count()%>%
  rename(`Count` = n) %>%
  mutate(Group = factor(Group, levels = c("CLAD-free", "CLAD"))) 

# Percentage of positive 
l <- list()
for (i in unique(dsa.det$Record.ID)){
  
  f <- dsa.det %>% 
    filter(Record.ID == i) %>%
    arrange(desc(Result))
  
  # If only one row
  if (dim(f)[1] < 2){
    # If Positive only
    if (f$Result %in% "1"){
      f <- f %>%
        mutate(Percentage_positive = round(100*(.$Count[1]/.$Count[1])))}
    # If Negative only
    if (f$Result %in% "0"){
      f <- f %>%
        mutate(Percentage_positive = round(100*(0/.$Count[1])))}}
  
  if (dim(f)[1] == 2){
    f <- f %>%
      mutate(Percentage_positive = round(100*(.$Count[2]/(.$Count[1] + .$Count[2]))))} 
  
  l[[i]] <- f}
dsa.det <- as.data.frame(bind_rows(l))

# Positive detected
name <- "dsa-detected"
p <- 
  
  dsa.det %>%
  filter(Result == "1") %>%
  ggplot(aes(x = Group, y = Percentage_positive)) +
  geom_boxplot(aes(fill = Group), outlier.shape = NA, alpha = 0.3) +
  geom_jitter(shape = 21, width = 0.2) +
  labs(title = "DSA detected",
       y = "Percentage positive per patient") +
  theme +
  theme(axis.title.x = element_blank()) +
  theme(legend.position = "bottom", legend.justification = c("left")) +
  stat_compare_means(label = "p.signif", 
                     method = "wilcox.test",
                     label.x.npc = "centre") +
  plot_fill(Group = TRUE)
ggsave(paste0("clinical-checks/", name, ".pdf"), p, width = 9, height = 7, units = 'cm') # ns

```



